---
title: "Shannon Diversity and Species Richess (rarefaction)"
author: "Kate"
date: "November 7, 2018"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width=5, fig.height=3, fig.align='center')
```


```{r packages and data, include=FALSE}

# Load necessary packages

library(tidyverse)
library(vegan)
library(VennDiagram)
library(eulerr)

# Load necessary data

smamms <- read_csv('./data/unique_smamms.csv') # using dataset with one row per unique individual
effort <- read_csv('./data/effort.csv')

```


```{r data wrangling, include=FALSE}

# Update classes

smamms <- smamms %>% 
  mutate(severity = factor(severity, levels = c('unb', 'mod', 'high')))

```


```{r species matrices, include=FALSE}

# Create a species matrix

matrix_temp <- smamms %>% 
  group_by(site, species) %>% 
  tally() %>% # count individuals per species per site
  spread(species, n) %>% # re-format into matrix
  ungroup() %>% 
  replace(., is.na(.), 0) # replace NAs with 0

sites <- matrix_temp$site # vector of just sites

matrix <- matrix_temp %>% 
  select(-site) # remove site column for input into vegan

matrix <- as.matrix(matrix)
rownames(matrix) <- sites



# Create a species matrix adjusted for trapping effort

matrix_temp <- smamms %>% 
  group_by(site, species) %>% 
  tally() %>% # count individuals per species per site
  full_join(., effort, by = 'site') %>% 
  mutate(n_adj = n/effort*300) %>% # add column for n adjusted for effort
  select(site, species, n_adj) %>% 
  spread(species, n_adj) %>% # re-format into matrix
  ungroup() %>% 
  replace(., is.na(.), 0) # replace NAs with 0

sites <- matrix_temp$site # vector of just sites

matrix_adj <- matrix_temp %>% 
  select(-site) # remove site column for input into vegan

matrix_adj <- as.matrix(matrix_adj)
rownames(matrix_adj) <- sites



# Create separate species matrices for each treatment

matrix_unb <- matrix[1:9,] # species matrix for low severity
matrix_mod <- matrix[19:27,] # species matrix for med severity
matrix_high <- matrix[10:18,] # species matrix for high severity

```

###Shannon Diversity

```{r calculate H, include=FALSE}

# Calculate Shannon-Weaver diversity index (H) for each site

H <- diversity(matrix_adj)

# Make a dataframe with site, severity, and H
diversity <- data_frame(sites, H) %>% 
  full_join(., smamms, by = c('sites' = 'site')) %>% 
  select(sites, severity, H) %>% 
  rename(site = sites) %>% 
  group_by(site) %>% 
  filter(row_number()==1)

```


```{r H data distribution, eval=FALSE, include=FALSE}

# Explore the data distribution

# Make histograms for each treatment
H_hist <- ggplot(diversity, aes(x = H)) +
  geom_histogram() +
  facet_wrap(~severity)

H_hist

# Make qqplots for each treatment
H_qq <- ggplot(diversity, aes(sample = H)) +
  geom_qq() +
  facet_wrap(~severity)

H_qq

# Histograms look a little funky because sample size is so small
# qq-plots for low and med severities look good, but qq-plot for high severity is skewed towards 0
# n=9 for each severity, so cannot use CLT to justify normality

```


```{r H data variance, eval=FALSE, include=FALSE}

# Explore variance
variances <- diversity %>% 
  group_by(severity) %>% 
  summarize(
    variance = var(H)
  )

variances

# The variances aren't that different for each severity

```


```{r H ANOVA and post-hoc tests, eval=FALSE, include=FALSE}

# Test for significant differences in means

# ANOVA
H_aov <- aov(H ~ severity, data = diversity)

summary(H_aov)
# p<0.05, so there's at least one sig dif in means

# Post-hoc pairwise testing

# Tukey's HSD
H_ph <- TukeyHSD(H_aov)
H_ph

# p=0.005 for low vs. high, but otherwise p>0.05
# significant difference between low and high severity means, but not between med vs. low or high severities

```


```{r H boxplot, echo=FALSE}

# Summarize data with boxplots
colors <- c('darkgreen', 'orange', 'red')
xlabs <- c('Unburned', 'Moderate Severity', 'High Severity') 

H_boxplot <- ggplot(diversity, aes(x = severity, y = H)) +
  geom_boxplot(aes(fill = severity)) +
  scale_fill_manual(values = colors) +
  theme_minimal() +
  labs(x = 'Fire Severity',
       y = 'Shannon Diversity (H)') +
  theme(legend.position = 'NA') +
  scale_x_discrete(labels = xlabs)

H_boxplot

# Calculate mean diversity for each treatment

H_summary <- diversity %>% 
  group_by(severity) %>% 
  summarise(
    H_mean = mean(H),
    H_sd = sd(H)
  )

```

**Shannon diversity (H) decreased across the fire severity gradient.** Diversity differed significantly between the three fire severity treatments (one-way ANOVA (2, 24) = 6.1, p < 0.01, n = 9 for each). Post-hoc analysis by Tukey’s HSD showed that mean diversity in unburned sites (H = 0.87) and high severity sites (H = 0.32) was significantly different (pairwise p = 0.005). 


###Species Richness (rarefaction)

```{r rarefaction, eval=FALSE, include=FALSE}

# Calculate richness for each site using rarefaction

raremax <- min(rowSums(matrix))
raremax # minimum sample count = 5: rarefy using this value

rare <- rarefy(matrix, sample = raremax)

# Make a dataframe with site, severity, and H

richness <- data_frame(sites, rare) %>% 
  full_join(., smamms, by = c('sites' = 'site')) %>% 
  select(sites, severity, rare) %>% 
  rename(site = sites) %>% 
  group_by(site) %>% 
  filter(row_number()==1)

```


```{r data distribution, eval=FALSE, include=FALSE}

# Explore the data distribution

# Make histograms for each treatment

rare_hist <- ggplot(richness, aes(x = rare)) +
  geom_histogram() +
  facet_wrap(~severity)

rare_hist

# Make qqplots for each treatment

rare_qq <- ggplot(richness, aes(sample = rare)) +
  geom_qq() +
  facet_wrap(~severity)

rare_qq

# Histograms look a little funky because sample size is so small
# qq-plots for low and med severities look good, but qq-plot for high severity is skewed towards 0
# n=9 for each severity, so cannot use CLT to justify normality

```


```{r data variance, eval=FALSE, include=FALSE}

# Explore variance

variances <- richness %>% 
  group_by(severity) %>% 
  summarize(
    variance = var(rare)
  )

variances

# The variances aren't that different for each severity

```


```{r ANOVA and post-hoc tests, eval=FALSE, include=FALSE}

# Test for significant differences in means

# ANOVA

rare_aov <- aov(rare ~ severity, data = richness)

summary(rare_aov)
# p<0.05, so there's at least one sig dif in means

# Post-hoc pairwise testing

# Tukey's HSD

rare_ph <- TukeyHSD(rare_aov)
rare_ph

# p=0.003 for low vs. high, but otherwise p>0.05
# significant difference between low and high severity means, but not between med vs. low or high severities

```


```{r rarefaction curves, fig.height=4, echo=FALSE}

# Rarefaction curves

accum_unb <- specaccum(matrix_unb, # unburned 
                       method = 'rarefaction', 
                       permutations = 100)

accum_mod <- specaccum(matrix_mod, # moderate severity
                       method = 'rarefaction',
                       permutations = 100)

accum_high <- specaccum(matrix_high, # high severity
                        method = 'rarefaction',
                        permutations = 100)

# Plot curves
plot(accum_unb,
     ci = 2, # 95% confidence interval
     ci.type = 'polygon',
     ci.col = rgb(red=0, green=1, blue=0, alpha=0.25),
     ci.lty = 0,
     lwd = 2,
     col = 'darkgreen',
     xlab = 'Sites',
     ylab = 'Rarefaction')
lines(accum_mod,
      ci = 2, # 95% confidence interval
      ci.type = 'polygon',
      ci.col = rgb(red=1, green=0.5, blue=0, alpha=0.25),
      ci.lty = 0,
      lwd = 2,
      col = 'orange')
lines(accum_high,
      ci = 2, # 95% confidence interval
      ci.type = 'polygon',
      ci.col = rgb(red=1, green=0, blue=0, alpha=0.25),
      ci.lty = 0,
      lwd = 2,
      col = 'red'
      )

```

**Species richness decreased across the fire severity gradient.** Individual-based rarefaction curves are plotted, with sites pooled for each severity. Differences in mean richness calculated from un-pooled rarefaction were significant (one-way ANOVA (2, 24) = 7.1, p < 0.01, n = 9 for each).

- No way to account for sampling effort (uses raw species matrix with counts)


```{r}

# Vectors of species at each severity

unb <- smamms %>% 
  filter(severity == 'unb')

mod <- smamms %>% 
  filter(severity == 'mod')

high <- smamms %>% 
  filter(severity == 'high')

unb_spp <- unique(unb$binomial)
mod_spp <- unique(mod$binomial)
high_spp <- unique(high$binomial)

all_spp <- list(unb_spp, mod_spp, high_spp)

venn.diagram(all_spp, filename = NULL)


euler()


```


