---
title: "Data Wrangling"
subtitle: "From [SMAMMUNITY: SMall mAMMal commUNITY](https://github.com/katekathrynkat/smammunity)"
output:
  html_document:
    theme: flatly
    highlight: tango
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = '..')
knitr::opts_chunk$set(include=FALSE)
```

**Kate Culhane**, Dept. of Ecology, Evolution, and Marine Biology, UC Santa Barbara  
kathrynculhane@ucsb.edu

**Publication:** Culhane, K., R. Sollmann, A.M. White, G.L. Tarbill, S. Cooper, M.A. Moritz, H.S. Young. (2021). Small mammal functional traits and vegetation characteristics mediate community responses to fire severity. Ecological Applications. Manuscript submitted for initial review.

**Source code:** [02_data_wrangling.Rmd](https://github.com/katekathrynkat/smammunity/blob/master/code/02_data_wrangling.Rmd)

---

### Pre-analysis manipulation of datasets

CSVs created for use in analyses (in folder "output_data"):

- **02_smamms.csv**: All small mammal capture data joined with site and species information

- **02_unique_smamms.csv**: Condensed small mammal capture data with one row per unique individual and no empty trap data

- **02_effort.csv**: Total trapping effort at each site (number of effective trap nights at each site during the entire trapping period, maximum of 300)

- **02_species_matrix.csv**: Species matrix for use in vegan

- **02_species_matrix_adj.csv**: Species matrix for use in vegan, adjusted for trapping effort

<br>

**Note:** These data sets INCLUDE rare species (species with five or less total captures). See [11. Analyses with rare species excluded](https://htmlpreview.github.io/?https://github.com/katekathrynkat/smammunity/blob/master/docs/11_rare_species_excluded.html) for analyses excluding these species. Rare species are:

- dusky-footed woodrat (NEFU)
- yellow pine chipmunk (NEAM)
- northern flying squirrel (GLSA)
- pinyon mouse (PETR)
- western harvest mouse (REME)

```{r packages and data}

# Packages
library(kableExtra) # making pretty tables
library(tidyverse)

# Small mammal capture data
smamms_clean <- read_csv('output_data/01_smamms_clean.csv')

# Site metadata
site_meta <- read_csv('raw_data/field_data/site_metadata.csv') %>%
    mutate(severity = factor(severity, levels = c('unb', 'mod', 'high')))

# Small mammal trait data
traits <- read_csv('raw_data/lit_data/mammal_traits.csv')

```

```{r}

# Create full smamm data frame with site and species information

smamms <- full_join(smamms_clean, site_meta) %>%
  left_join(traits) %>% 
  select(site, severity, date, day, trap, indID, species, binomial, binomial_short, common_name, family, age, recap, weight, comments)

write_csv(smamms, 'output_data/02_smamms.csv', col_names = TRUE)

```

```{r}

# Create data frame with one row per unique individual

unique_smamms <- smamms %>% 
  filter(species != 'MISS' & species != 'SPRU') %>% # remove SPRU and MISS
  group_by(indID) %>% 
  arrange(-day) %>% 
  filter(row_number()==1) %>% # filter by unique animals (only keep record from last capture)
  ungroup() %>% 
  select(-day)
    
write_csv(unique_smamms, 'output_data/02_unique_smamms.csv', col_names = TRUE)

```

```{r}

# Data frame of trapping effort at each site

effort <- smamms %>% 
  filter(species == 'SPRU' | species == 'MISS') %>%
  mutate(usage = case_when(
    species == 'SPRU' ~ 0.5, # sprung traps coded as 0.5 usage
    species == 'MISS' ~ 1 # missing traps coded as 1 usage (to be subtracted)
  )) %>% 
  group_by(site, severity) %>% 
  summarize(sprung = sum(usage)) %>% 
  mutate(effort =300 - sprung) %>% 
  select(site, severity, effort) %>% 
  arrange(severity)

write_csv(effort, 'output_data/02_effort.csv', col_names = TRUE)

```

```{r}

# Species matrix for use in vegan

matrix <- unique_smamms %>% 
  group_by(site, severity, species) %>% 
  tally() %>% # count individuals per species per site
  spread(species, n) %>% # re-format into matrix
  ungroup() %>% 
  replace(., is.na(.), 0) %>% # replace NAs with 0
  arrange(severity)

write_csv(matrix, 'output_data/02_species_matrix.csv')

# Species matrix adjusted for trapping effort

matrix_adj <- unique_smamms %>%
  group_by(site, severity, species) %>%
  tally() %>% # count individuals per species per site
  full_join(effort, by = c('site', 'severity')) %>%
  mutate(n_adj = n/effort*300) %>% # add column for n adjusted for effort
  select(site, species, n_adj)%>%
  spread(species, n_adj) %>% # re-format into matrix
  ungroup() %>%
  replace(., is.na(.), 0) %>% # replace NAs with 0
  arrange(severity)

write_csv(matrix_adj, 'output_data/02_species_matrix_adj.csv')

```

---

**Site by species matrix**

```{r, include=TRUE, echo=FALSE}

kable(matrix) %>% 
  kable_styling(bootstrap_options = c('striped', 'condensed'),
                full_width = FALSE,
                position = 'left')

```

```{r, eval=FALSE}
rmarkdown::render('code/02_data_wrangling.Rmd', output_file = '../docs/02_data_wrangling.html')
```
