---
title: "Small Mammal Richness and Evenness"
subtitle: "From [SMAMMUNITY: SMall mAMMal commUNITY](https://github.com/katekathrynkat/smammunity)"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = '..')
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.width=5, fig.height=3)
```

**Kate Culhane**, Dept. of Ecology, Evolution, and Marine Biology, UC Santa Barbara\
[kathrynculhane\@ucsb.edu](mailto:kathrynculhane@ucsb.edu){.email}

**Publication:** Culhane, K., R. Sollmann, A.M. White, G.L. Tarbill, S. Cooper, M.A. Moritz, H.S. Young. (2021). Small mammal functional traits and vegetation characteristics mediate community responses to fire severity. Ecological Applications. Manuscript submitted for initial review.

**Source code:** [06_diversity.Rmd](https://github.com/katekathrynkat/smammunity/blob/master/code/06_diversity.Rmd)

---

```{r packages and data}

# Packages
library(vegan) # community analyses
library(iNEXT) # rarefaction curves
library(dunn.test) # Dunn's test (KW post hoc)
library(rstatix) # 
library(ggpubr) # 
library(effectsize) # for eta-squared
library(tidyverse)

# Small mammal site by species matrix
matrix_df <- read_csv('output_data/02_species_matrix.csv') %>% 
  mutate(severity = factor(severity, levels = c('unb', 'mod', 'high')))

# Small mammal site by species matrix adjusted for trapping effort
matrix_adj_df <- read_csv('output_data/02_species_matrix_adj.csv') %>%
  mutate(severity = factor(severity, levels = c('unb', 'mod', 'high')))

# Small mammal trapping effort
effort <- read_csv('output_data/02_effort.csv')

```

```{r data wrangling}

# Transform data frames into matrices

matrix <- as.matrix(matrix_df[-c(1,2)])
rownames(matrix) <- matrix_df$site

matrix_adj <- as.matrix(matrix_adj_df[-c(1,2)])
rownames(matrix_adj) <- matrix_adj_df$site

# Create separate species matrices for each treatment

matrix_unb <- matrix[1:9,] # species matrix for low severity
matrix_mod <- matrix[10:18,] # species matrix for med severity
matrix_high <- matrix[19:27,] # species matrix for high severity

# Create matrix with species pooled for each treatment

matrix_pooled_df <- matrix_df %>% 
  group_by(severity) %>% 
  summarize_if(is.numeric, sum)

matrix_pooled <- as.matrix(matrix_pooled_df[-1])
rownames(matrix_pooled) <- matrix_pooled_df$severity

# Vectors for boxplots

cols <- c('darkgreen', 'orange', 'red')
xlabs <- c('Unburned', 'Moderate Severity', 'High Severity') 

```

## Species Richness

### Number of species

**Number of species per site, by fire severity:**

```{r}

# Calculate species richness (number of species) for each site
n <- specnumber(matrix)

# Make a dataframe with site, severity, and number of species
n_spp <- tibble(site = matrix_df$site, severity = matrix_df$severity, n_spp = n)

# Jitterplot with mean/SE
ggplot(n_spp, aes(x = severity, y = n)) +
  geom_jitter(aes(color = severity)) +
  scale_color_manual(values = cols) +
  stat_summary(fun.data = mean_se, geom = "crossbar", width = .5) +
  theme_classic() +
  labs(x = 'Fire Severity',
       y = 'Number of Species') +
  theme(legend.position = 'NA') +
  scale_x_discrete(labels = xlabs)

```

- Nine species were found at unburned sites, eight at low-moderate severity sites, and seven at high severity sites
- Shared species: *P. maniculatus*, *N. quadrimaculatus*, *O. beecheyi*, *P. boylii*
- Species unique to unburned sites: *G. sabrinus*, *R. megalotis*

<br>

### Individual-based rarefaction

```{r}

# Rarefaction
accum_unb <- specaccum(matrix_unb, method = 'rarefaction', permutations = 1000)
accum_mod <- specaccum(matrix_mod, method = 'rarefaction', permutations = 1000)
accum_high <- specaccum(matrix_high, method = 'rarefaction', permutations = 1000)

# ggplot using individual-based rarefaction

unb_data <- data.frame(site = accum_unb$sites,
                       indv = accum_unb$individuals,
                       richness = accum_unb$richness,
                       CI = accum_unb$sd)
mod_data <- data.frame(site = accum_mod$sites,
                       indv = accum_mod$individuals,
                       richness = accum_mod$richness,
                       CI = accum_mod$sd)
high_data <- data.frame(site = accum_high$sites,
                        indv = accum_high$individuals,
                        richness = accum_high$richness,
                        CI = accum_high$sd)

ggplot() +
  geom_line(data = unb_data,
            aes(x = indv, y = richness),
            size = 1, color = 'darkgreen') +
  geom_ribbon(data = unb_data,
              aes(x = indv, ymin = (richness-2*CI), ymax=(richness+2*CI)),
              alpha = 0.2, fill = 'darkgreen') +
  geom_line(data = mod_data,
            aes(x = indv, y = richness),
            size = 1, color = 'orange') +  
  geom_ribbon(data = mod_data,
              aes(x = indv, ymin = (richness-2*CI), ymax=(richness+2*CI)),
              alpha = 0.2, fill = 'orange') +
  geom_line(data = high_data,
            aes(x = indv, y = richness),
            size = 1, color = 'red') +
  geom_ribbon(data = high_data,
              aes(x = indv, ymin = (richness-2*CI), ymax=(richness+2*CI)),
              alpha = 0.2, fill = 'red') +
  scale_y_continuous(breaks = c(2,4,6,8,10)) +
  labs(x = 'Individuals Caught',
       y = 'Richness') +
  theme_classic() +
  theme(axis.title.x = element_text(size = 13),
        axis.title.y = element_text(size = 13),
        axis.text.x = element_text(size = 11, color = 'black'),
        axis.text.y = element_text(size = 11, color = 'black'),
        plot.margin = unit(c(10,10,10,10), 'pt'))

```

- Sites pooled by severity
- Scaled to individuals

<br>

```{r, fig.width=7}

# Individual based rarefaction

inext_matrix <- t(matrix_pooled)

out <- iNEXT(inext_matrix,
             q = 0, # the Hill number to estimate (q = 0 gives richness)
             datatype = 'abundance') # vs. incidence (presence/absence data)

# Plot curves

df <- fortify(out, type=1) %>% 
  mutate(site = factor(site, levels = c('unb', 'mod', 'high')))

df <- fortify(out, type=1) %>% 
  mutate(site = factor(site, levels = c('unb', 'mod', 'high')))
df.point <- df[which(df$method=='observed'),]
df.line <- df[which(df$method!='observed'),]
df.line$method <- factor(df.line$method, 
                         c('interpolated', 'extrapolated'),
                         c('interpolation', 'extrapolation'))
 
ggplot(df, aes(x = x, y = y, color = site)) + 
  geom_line(data = df.line, aes(linetype = method),
            lwd = 1) +
  geom_ribbon(aes(ymin = y.lwr, ymax = y.upr, fill = site, color = NULL),
              alpha=0.2) +
  scale_linetype(name = 'Method', labels = c('Interpolation', 'Extrapolation')) +
  scale_color_manual(name = 'Fire severity', values = c('#4AB793', '#DCA827', '#D85F2B'),
                     labels = c('Unburned', 'Low-moderate severity', 'High severity')) +
  scale_fill_manual(name = 'Fire severity', values = c('darkgreen', 'orange', 'red'),
                    labels = c('Unburned', 'Low-moderate severity', 'High severity')) +
  scale_y_continuous(breaks = c(2,4,6,8,10,12)) +
  labs(x = 'Number of individuals',
       y = 'Number of species') +
  guides(color = guide_legend(order=1),
         fill = guide_legend(order=1),
         shape = guide_legend(order=2)) +
  theme(legend.title = element_blank()) +
  theme_classic()

ggsave('figs/06_figS1.png', plot = last_plot(),
       width = 6, height = 3, units = 'in')

```

- **Supplementary Figure 1**
- Sites pooled by severity
- Scaled to individuals
- Unconstrained 95% confidence intervals

<br>

### Per-site richness

(individual-based rarefaction estimates)

```{r rarefaction, include=FALSE}

# Calculate richness for each site using rarefaction
raremax <- min(rowSums(matrix))
raremax # minimum sample count = 5: rarefy using this value
rarefied_richness <- rarefy(matrix, sample = raremax)

# Make a dataframe with site, severity, and richness
richness <- tibble(site = matrix_df$site, severity = matrix_df$severity, rarefied_richness)

```

```{r data distribution, eval=FALSE}

# Check summary stats and model assumptions for richness

# Summary table
richness %>% 
  group_by(severity) %>% 
  get_summary_stats(rarefied_richness)

# Check for outliers
richness %>% group_by(severity) %>% 
  identify_outliers(rarefied_richness)
# none

# Check for normality
lm <- lm(rarefied_richness~severity, data=richness)
ggqqplot(residuals(lm))
ggqqplot(richness, "rarefied_richness", facet.by="severity")
shapiro_test(residuals(lm))
richness %>% group_by(severity) %>% 
  shapiro_test(rarefied_richness)
# Looks pretty good

# Homogoneity of variance
plot(lm,1)
richness %>% levene_test(rarefied_richness~severity)
# Variance is fine

```

```{r Rarefied Richness boxplot}

# Jitterplot with mean/SE
ggplot(richness, aes(x = severity, y = rarefied_richness)) +
  geom_jitter(aes(color = severity)) +
  scale_color_manual(values = cols) +
  stat_summary(fun.data = mean_se, geom = "crossbar", width = .5) +
  theme_classic() +
  labs(x = 'Fire Severity',
       y = 'Rarefied Richness') +
  theme(legend.position = 'NA') +
  scale_x_discrete(labels = xlabs)

# Stats
aov <- aov(rarefied_richness~severity, data=richness)
summary(aov)
richness %>% ungroup() %>% tukey_hsd(rarefied_richness~severity)
eta_squared(aov)

#Kruskal-Wallis
kruskal_test(rarefied_richness~severity, data=richness)
kruskal_effsize(rarefied_richness~severity, data=richness)
dunn.test(richness$rarefied_richness, richness$severity, method='bonferroni')

```

- Significantly different by ANOVA

<br>

## Pielou's Evenness (J)

```{r}

# Calculate Pielou's Evenness

J <- diversity(matrix_adj)/log(specnumber(matrix_adj)) # H / log (number of species)

diversity <- cbind(richness, J)
diversity$J[is.nan(diversity$J)] <- NA

```

**Pielou's Evenness (J) decreased across the fire severity gradient:**

```{r J boxplot}

# Jitterplot with mean/SE
ggplot(diversity, aes(x = severity, y = J)) +
  geom_jitter(aes(color = severity)) +
  scale_color_manual(values = cols) +
  stat_summary(fun.data = mean_se, geom = "crossbar", width = .5) +
  theme_classic() +
  labs(x = 'Fire Severity',
       y = 'Pielou\'s Evenness (J)') +
  theme(legend.position = 'NA') +
  scale_x_discrete(labels = xlabs)

```

```{r J data distribution, eval=FALSE}

# Check summary stats and model assumptions for evenness

# Summary table
diversity %>% 
  group_by(severity) %>% 
  get_summary_stats(J)

# Check for outliers
diversity %>% group_by(severity) %>% 
  identify_outliers(J)
# 3 outliers (GRNCAR, GRNKAN, GRNWOO)

# Check for normality
lm <- lm(J~severity, data=diversity)
ggqqplot(residuals(lm))
ggqqplot(diversity, "J", facet.by="severity")
shapiro_test(residuals(lm))
diversity %>% group_by(severity) %>% 
  shapiro_test(J)
# Looks pretty good

# Homogoneity of variance
plot(lm,1)
diversity %>% levene_test(J~severity)
# Variance is fine

```

```{r}

# Stats
aov <- aov(J~severity, data=diversity)
summary(aov)
diversity %>% ungroup() %>% tukey_hsd(J~severity)
cohens_f(aov)

#Kruskal-Wallis
kruskal_test(J~severity, data=diversity)
kruskal_effsize(J~severity, data=diversity)
dunn.test(diversity$J, diversity$severity, method='bonferroni')

```

- Significantly different by ANOVA

<br>

## Shannon Diversity (H)

```{r calculate H}

# Calculate Shannon-Weaver diversity index (H) for each site (adjusted for trapping effort)

H <- diversity(matrix_adj)

diversity <- cbind(diversity, H)

# Export all diversity data

write_csv(diversity, 'output_data/06_diversity.csv', col_names = TRUE)

```

**Shannon diversity (H) decreased across the fire severity gradient:**

```{r H boxplot}

# Jitterplot with mean/SE
ggplot(diversity, aes(x = severity, y = H)) +
  geom_jitter(aes(color = severity)) +
  scale_color_manual(values = cols) +
  stat_summary(fun.data = mean_se, geom = "crossbar", width = .5) +
  theme_classic() +
  labs(x = 'Fire Severity',
       y = 'Shannon Diversity (H)') +
  theme(legend.position = 'NA') +
  scale_x_discrete(labels = xlabs)

```

```{r H data distribution, eval=FALSE}

# Check summary stats and model assumptions for evenness

# Summary table
diversity %>% 
  group_by(severity) %>% 
  get_summary_stats(H)

# Check for outliers
diversity %>% group_by(severity) %>% 
  identify_outliers(H)

# Check for normality
lm <- lm(H~severity, data=diversity)
ggqqplot(residuals(lm))
ggqqplot(diversity, "H", facet.by="severity")
shapiro_test(residuals(lm))
diversity %>% group_by(severity) %>% 
  shapiro_test(H)
# Looks pretty good

# Homogoneity of variance
plot(lm,1)
diversity %>% levene_test(H~severity)
# Variance is fine

```

```{r}

# Stats

aov <- aov(H~severity, data=diversity)
summary(aov)
diversity %>% ungroup() %>% tukey_hsd(H~severity)
cohens_f(aov)

```

- Significantly different by ANOVA

```{r eval=FALSE}
rmarkdown::render('code/06_diversity.Rmd', output_file = '../docs/06_diversity.html')
```
