---
title: "Spatial Metrics"
author: "Kate Culhane"
date: "January 7, 2019"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = '..')
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

<br>

### Spatial Data

- **Monitoring Trends in Burn Severity (MTBS)**
  - Fire severity data
- **LANDFIRE**
  - Data from 2012 (pre-burn)
    - VCC
    - EVC
    - EVT
  - Data from 2014 (post-burn)
    - EVC
    - EVT

```{r packages and data, include=FALSE}

# Load necessary packages

library(tidyverse)
library(raster) # for raster work
library(rgdal) # for vector work (includes 'sp')
library(sf)
library(ggspatial)
library(mapview)
library(spatialEco) # for several landscape metrics
library(geosphere)
library(rgeos) # gDistance function

# Update functions

select <- dplyr::select
summarize <- dplyr::summarize
rename <- dplyr::rename

# Load necessary data

site_meta <- read_csv('raw_data/field_data/site_metadata.csv') %>% 
  mutate(severity = factor(severity, levels = c('unb', 'mod', 'high')))

# Load spatial data

sites_dat <- read_csv('raw_data/spatial_data/site_coords.csv')

# MTBS data

mtbs_info <- GDALinfo('raw_data/spatial_data/ca3878212060420140913_20130730_20150805_dnbr6.tif') # metadata
mtbs_raster <- raster('raw_data/spatial_data/ca3878212060420140913_20130730_20150805_dnbr6.tif') # burn severity
bndy_shp <- readOGR(dsn = 'raw_data/spatial_data',
                    layer = 'ca3878212060420140913_20130730_20150805_burn_bndy') # burn boundary

# LANDFIRE data

vvc_landfire <- raster("raw_data/spatial_data/vcc/hdr.adf") # vegetation condition classes
vcc_raster <- projectRaster(vvc_landfire, crs=crs(mtbs_raster))

evt_landfire <- raster("raw_data/spatial_data/evt_2012/hdr.adf") # existing vegetation type
evt_raster <- projectRaster(evt_landfire, crs=crs(mtbs_raster))
evc_landfire <- raster("raw_data/spatial_data/evc_2012/hdr.adf") # existing vegetation cover
evc_raster <- projectRaster(evc_landfire, crs=crs(mtbs_raster))

evt_landfire2 <- raster("raw_data/spatial_data/evt_2014/hdr.adf") # existing vegetation type
evt_raster2 <- projectRaster(evt_landfire2, crs=crs(mtbs_raster))
evc_landfire2 <- raster("raw_data/spatial_data/evc_2014/hdr.adf") # existing vegetation cover
evc_raster2 <- projectRaster(evc_landfire2, crs=crs(mtbs_raster))

```

### Burn severity map

```{r spatial data wrangling, include=FALSE}

# Vector shapefile for the site corners

corners_dat <- sites_dat %>% 
  filter(type == 'Plot Corner') %>% 
  mutate(northing = northing + 10000000)

corners_points <- corners_dat

coordinates(corners_points) <- ~ easting + northing

proj4string(corners_points) <- '+proj=utm +zone=10 +south +ellps=WGS84 +datum=WGS84 +units=m +no_defs' # CRS: WGS84 / UTM zone 10S

crs(mtbs_raster) # CRS for MTBS raster is NAD83 / AEA (Albers Equal-Area Conic)

corners_aea <- spTransform(corners_points, CRS('+proj=aea +lat_1=29.5 +lat_2=45.5 +lat_0=23 +lon_0=-96 +x_0=0 +y_0=0
+datum=NAD83 +units=m +no_defs +ellps=GRS80 +towgs84=0,0,0 ')) # update site CRS to match mtbs CRS

# Vector shapefile for the site centroids

centers_dat <- corners_dat %>% 
  group_by(site) %>% 
  summarize(x = mean(easting), # average corner coords for centroid coords
            y = mean(northing))

site_points <- centers_dat

coordinates(site_points) <- ~ x + y

proj4string(site_points) <- '+proj=utm +zone=10 +south +ellps=WGS84 +datum=WGS84 +units=m +no_defs' # CRS: WGS84 / UTM zone 10S

sites_aea <- spTransform(site_points, CRS('+proj=aea +lat_1=29.5 +lat_2=45.5 +lat_0=23 +lon_0=-96 +x_0=0 +y_0=0
+datum=NAD83 +units=m +no_defs +ellps=GRS80 +towgs84=0,0,0 ')) # update site CRS to match mtbs CRS

# Export to csv

x <- coordinates(sites_aea)[,1]
y <- coordinates(sites_aea)[,2]

site_centroids <- data.frame(sites_aea$site, x, y)

write_csv(site_centroids,'output_data/site_centroids.csv', col_names = TRUE)

```

```{r create site polygons, include=FALSE}

# Create polygons from site corner coords

corners_df <- as.data.frame(corners_aea) %>% 
  mutate(order = case_when(
    point == 'N' ~ 1, point == 'W' ~ 2, point == 'S' ~ 3, point == 'E' ~ 4,
    point == 'NW' ~ 1, point == 'NE' ~ 2, point == 'SE' ~ 3, point == 'SW' ~ 4
  )) %>% 
  arrange(desc(order))# update order of points so sites are rectangles, not bowties

sites_list <- list() # empty list to fill with polys for each site

for (i in unique(corners_df$site)) {
  df <- corners_df %>% 
    filter(site == i) %>% # filter for each site 
    select(easting, northing) %>%
    rename(x = easting, y = northing)
  matrix <- base::as.matrix(df) # matrix with 1 col each for x, y
  poly <- Polygon(matrix) # create polygon object
  poly2 <- Polygons(list(poly), 1) # create multi-ring polygon object
  poly3 <- SpatialPolygons(list(poly2)) # create SpatialPolygon objecct
  proj4string(poly3) <-'+proj=aea +lat_1=29.5 +lat_2=45.5 +lat_0=23 +lon_0=-96 +x_0=0 +y_0=0 +datum=NAD83 +units=m +no_defs +ellps=GRS80 +towgs84=0,0,0 ' # update CRS
  assign(paste0(i, '_poly'), poly3) # assign each polygon a name
  sites_list[[i]] <- poly3 # add each SpatialPolygons object to a list
}

```

```{r}

# Create polygons from mammal grid corner coords

grid_corners <- sites_dat %>% 
  filter(type == 'Mammal Grid') %>% 
  mutate(northing = northing + 10000000)

coordinates(grid_corners) <- ~ easting + northing

proj4string(grid_corners) <- '+proj=utm +zone=10 +south +ellps=WGS84 +datum=WGS84 +units=m +no_defs' # CRS: WGS84 / UTM zone 10S

grid_corners_aea <- spTransform(grid_corners, CRS('+proj=aea +lat_1=29.5 +lat_2=45.5 +lat_0=23 +lon_0=-96 +x_0=0 +y_0=0
+datum=NAD83 +units=m +no_defs +ellps=GRS80 +towgs84=0,0,0 ')) # update site CRS to match mtbs CRS

grid_corners_df <- as.data.frame(grid_corners_aea) %>% 
  mutate(order = case_when(
    point == 'A1' ~ 1,
    point == 'A10' ~ 2, 
    point == 'J10' ~ 3,
    point == 'J1' ~ 4
  )) %>% 
  arrange(desc(order))# update order of points so that sites are rectangles instead of bowties

grids_list <- list() # empty list to fill with polys for each site

for (i in unique(grid_corners_df$site)) {
  df <- grid_corners_df %>% 
    filter(site == i) %>% # filter for each site 
    select(easting, northing) %>%
    dplyr::rename(x = easting, y = northing)
  matrix <- base::as.matrix(df) # matrix with 1 col each for x, y
  poly <- Polygon(matrix) # create polygon object
  poly2 <- Polygons(list(poly), 1) # create multi-ring polygon object
  poly3 <- SpatialPolygons(list(poly2)) # create SpatialPolygon objecct
  proj4string(poly3) <-'+proj=aea +lat_1=29.5 +lat_2=45.5 +lat_0=23 +lon_0=-96 +x_0=0 +y_0=0 +datum=NAD83 +units=m +no_defs +ellps=GRS80 +towgs84=0,0,0 ' # update CRS
  assign(paste0(i, '_poly'), poly3) # assign each polygon a name
  grids_list[[i]] <- poly3 # add each SpatialPolygons object to a list
}

# Largest homerange (GLSA) = 0.04 km^2 = 40000 m^2
# Area of a circle = pi * r^2
# 40000 = pi * r^2
# r = sqrt(40000/pi) = 112.8379 m
# d = 2r = 225.6758 m

# Create polygons of each grid + buffer of 112.8379m

buffsites_list <- list() # empty list to fill with buffered polys

for (i in unique(grid_corners_df$site)) {
  poly <- spTransform(grids_list[[i]], CRS('+proj=utm +zone=10 +south +ellps=WGS84 +datum=WGS84 +units=m +no_defs')) # transform to WGS to work with units of meters
  buffpoly <- buffer(poly, width = 225.6758) # create buffered polygon
  buffpoly_aea <- spTransform(buffpoly, CRS('+proj=aea +lat_1=29.5 +lat_2=45.5 +lat_0=23 +lon_0=-96 +x_0=0 +y_0=0
+datum=NAD83 +units=m +no_defs +ellps=GRS80 +towgs84=0,0,0 ')) # transform back to AEA/NAD83
  assign(paste0(i, '_buffpoly'), buffpoly_aea) # assign each polygon a name
  buffsites_list[[i]] <- buffpoly_aea # add each SpatialPolygons object to a list
}

plot.new()
plot(mtbs_raster)
for (i in unique(grid_corners_df$site)) {
  plot(buffsites_list[[i]],
       col = 'blue',
       add = TRUE)
} # looks good!

```

- Burn severity data from MTBS (Monitoring Trends in Burn Severity)
- Buffered mammal grids shown in blue (mammal grids buffered by the largest home range of a trapped mammals)

<br>

### Pre-disturbance habitat characteristics

#### Existing Vegetation Type (EVT)

```{r}

# Map of EVT in the northern burn area

plot.new()
plot(evt_raster)

```

- **Average vegetation type:** average LANDFIRE EVT pixel value within buffered grid boundaries

```{r evt within expanded sites, eval=FALSE}

# EVT 2012

evt_pixels <- data.frame(site = character(), # blank data frame
                         evt = numeric())

for (i in unique(grid_corners_df$site)) {
  df <- extract(evt_raster, buffsites_list[[i]], df = TRUE) %>% # extract pixel info
    mutate(site = i) %>% 
    dplyr::rename(evt = hdr) %>% 
    select(site, evt)
  evt_pixels <- rbind(evt_pixels, df) # add pixel information to df, one row per pixel
}

evt_summary <- evt_pixels %>% 
  full_join(site_meta[3:4]) %>% 
  mutate(evt = round(evt,0)) %>% 
  group_by(site, severity) %>% 
  count(evt)
  
write_csv(evt_summary, 'output_data/EVT_2012.csv', col_names = TRUE)

# EVT 2014

evt_pixels2 <- data.frame(site = character(), # blank data frame
                         evt = numeric())

for (i in unique(grid_corners_df$site)) {
  df <- extract(evt_raster2, buffsites_list[[i]], df = TRUE) %>% # extract pixel info
    mutate(site = i) %>% 
    dplyr::rename(evt = hdr) %>% 
    select(site, evt)
  evt_pixels2 <- rbind(evt_pixels2, df) # add pixel information to df, one row per pixel
}

evt_summary2 <- evt_pixels2 %>% 
  full_join(site_meta[3:4]) %>% 
  mutate(evt = round(evt,0)) %>% 
  group_by(site, severity) %>% 
  count(evt)
  
write_csv(evt_summary2, 'output_data/EVT_2014.csv', col_names = TRUE)

```

#### Existing Vegetation Cover (EVC)

```{r}

# Map of EVC in the northern burn area

plot.new()
plot(evc_raster)

```

- **Average vegetation cover:** average LANDFIRE EVC pixel value within buffered grid boundaries

```{r evt within expanded sites, eval=FALSE}

# EVC 2012

evc_pixels <- data.frame(site = character(), # blank data frame
                         evc = numeric())

for (i in unique(grid_corners_df$site)) {
  df <- extract(evc_raster, buffsites_list[[i]], df = TRUE) %>% # extract pixel info
    mutate(site = i) %>% 
    dplyr::rename(evc = hdr) %>% 
    select(site, evc)
  evc_pixels <- rbind(evc_pixels, df) # add pixel information to df, one row per pixel
}

evc_summary <- evc_pixels %>% 
  full_join(site_meta[3:4]) %>% 
  mutate(evc = round(evc,0)) %>% 
  group_by(site, severity) %>% 
  count(evc) 

write_csv(evc_summary, 'output_data/EVC_2012.csv', col_names = TRUE)

# EVC 2012

evc_pixels2 <- data.frame(site = character(), # blank data frame
                         evc = numeric())

for (i in unique(grid_corners_df$site)) {
  df <- extract(evc_raster2, buffsites_list[[i]], df = TRUE) %>% # extract pixel info
    mutate(site = i) %>% 
    dplyr::rename(evc = hdr) %>% 
    select(site, evc)
  evc_pixels2 <- rbind(evc_pixels2, df) # add pixel information to df, one row per pixel
}

evc_summary2 <- evc_pixels2 %>% 
  full_join(site_meta[3:4]) %>% 
  mutate(evc = round(evc,0)) %>% 
  group_by(site, severity) %>% 
  count(evc) 

write_csv(evc_summary2, 'output_data/EVC_2014.csv', col_names = TRUE)

```


#### Vegetation Condition Classes (VCC)

```{r}

# Map of VCC in the northern burn area

plot.new()
plot(vcc_raster)

```

- **Average vegetation cover:** average LANDFIRE VCC pixel value within buffered grid boundaries

```{r evt within expanded sites, eval=FALSE}

# Create data frame with a row for each pixel in each expanded site

vcc_pixels <- data.frame(site = character(), # blank data frame
                         vcc = numeric())

for (i in unique(grid_corners_df$site)) {
  df <- extract(vcc_raster, buffsites_list[[i]], df = TRUE) %>% # extract pixel info
    mutate(site = i) %>% 
    dplyr::rename(vcc = hdr) %>% 
    select(site, vcc)
  vcc_pixels <- rbind(vcc_pixels, df) # add pixel information to df, one row per pixel
}

vcc_summary <- vcc_pixels %>% 
  full_join(site_meta[3:4]) %>% 
  mutate(vcc = round(vcc,0)) %>% 
  group_by(site, severity) %>% 
  count(vcc)%>% 
  filter(n/sum(n) > 0.1) %>% # remove pixels that make up <10% of the habitat 
  mutate(vcc = case_when(
    vcc == 3 ~ 'Vegetation Condition Class II.A Moderate to Low, Vegetation Departure 34-50% Vegetation Condition Class II Moderate Vegetation Departure',
    vcc == 4 ~ 'Vegetation Condition Class II.B Moderate to High, Vegetation Departure 51-66% Vegetation Condition Class II Moderate Vegetation Departure'
  ))

```

<br>

### Calculating spatial metrics

- **Average severity:** average MTBS pixel value within buffered site boundaries

```{r avg severity within expanded sites, eval=FALSE}

# Create data frame with a row for each pixel in each expanded site

buffsite_pixels <- data.frame(site = character(), # blank data frame
                           severity = numeric())

for (i in unique(grid_corners_df$site)) {
  df <- extract(mtbs_raster, buffsites_list[[i]], df = TRUE) %>% # extract pixel information within each buffered site polygon
    mutate(site = i) %>% 
    dplyr::rename(severity = ca3878212060420140913_20130730_20150805_dnbr6) %>% 
    select(site, severity)
  buffsite_pixels <- rbind(buffsite_pixels, df) # add pixel information to df, one row per pixel
}

# Summary of % of each severity within each buffered site

buffpixel_summary <- buffsite_pixels %>% 
  group_by(site, severity) %>% 
  summarize(pixels = length(severity)) %>% 
  group_by(site) %>% 
  mutate(prop = pixels/sum(pixels))

# Calculate average severity value for each site

buffsite_sev_avg <- buffsite_pixels %>% 
  group_by(site) %>% 
  summarize(avg = mean(severity))

```

- **Distance to burn boundary:** distance from each site centroid to the nearest burn boundary (m)

```{r calculate distance to burn boundary, eval=FALSE}

# Calculate distance to burn boundary

burn_edge <- as(bndy_shp, 'SpatialLinesDataFrame')

distances <- t(gDistance(sites_aea, burn_edge, byid = TRUE))

distance_to_edge <- data.frame(sites_aea$site, distances) %>% 
  rename(site = sites_aea.site) %>% 
  full_join(site_meta) %>% 
  mutate(distance = case_when(
    severity == 'unb' ~ -X0,
    TRUE ~ X0
  ))

```

```{r, eval=FALSE}

# Export data

spatial_metrics <- full_join(buffsite_sev_avg, distance_to_edge) %>% 
  select(site, severity, avg, distance) %>% 
  rename(avg_pixel = avg,
         dist_to_edge = distance)

write_csv(spatial_metrics, 'output_data/spatial_metrics.csv', col_names = TRUE)

```

```{r distance between sites, eval=FALSE}

# Calculate distances between each pair of sites

dist_all <- as.data.frame(pointDistance(site_points))

dist_unb <- as.data.frame(pointDistance(site_points[1:9,]))

dist_mod <- as.data.frame(pointDistance(sites_aea[19:27,]))

dist_high <- as.data.frame(pointDistance(sites_aea[10:18,]))

```

```{r eval=FALSE}

rmarkdown::render('code/06_spatial_metrics.Rmd', output_file = '../docs/06_spatial_metrics.html')

```
