---
title: "Habitat Metric PCA"
author: "Kate Culhane"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = '..')
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.width=5, fig.height=3)
```

<br>

```{r packages and data}

# Load packages

library(tidyverse)
library(ggbiplot)
library(randomForest)
library(mvabund)
library(corrplot)

# Update functions

rename <- dplyr::rename
select <- dplyr::select

# Load data

matrix_df <- read_csv('output_data/species_matrix.csv') %>% 
  mutate(severity = factor(severity, levels = c('unb', 'mod', 'high')))

matrix_adj_df <- read_csv('output_data/species_matrix_adj.csv') %>% 
  mutate(severity = factor(severity, levels = c('unb', 'mod', 'high')))

site_meta <- read_csv('raw_data/field_data/site_metadata.csv') %>% 
  mutate(severity = factor(severity, levels = c('unb', 'mod', 'high')))

veg <- read_csv('output_data/veg_metrics.csv')

spatial <- read_csv('output_data/spatial_metrics.csv')

diversity <- read_csv('output_data/diversity.csv') %>% 
  mutate(severity = factor(severity, levels = c('unb', 'mod', 'high')))

```


```{r data wrangling}

# Data for diversity GLM

habitat <- full_join(spatial, veg) %>% 
  mutate(severity = factor(severity, levels = c('unb', 'mod', 'high'))) %>% 
  arrange(severity)

```


### Correlated habitat variables

**Correlation plot of habitat variables:**

```{r correlation plot, fig.height=5}

# Create a correlation matrix

habitat_cor <- habitat %>% 
  select(-site, -severity) %>% # select only continuous variables
  cor(.) # create correlation matrix

# Visualize the correlation matrix

corrplot(habitat_cor,
         type = 'upper',
         order = 'hclust', # order variables by cluster
         tl.col = 'black',
         tl.cex = 0.8)

# Rank correlations

as.data.frame(habitat_cor) %>% 
  mutate(variable1 = rownames(habitat_cor)) %>%
  gather(variable2, cor_strength, 1:14) %>% 
  filter(cor_strength != 1,
         cor_strength > 0.7 | cor_strength < -0.7) %>% 
  arrange(cor_strength) %>% 
  distinct(cor_strength, .keep_all = TRUE)

```

**Highly correlated:**

- All fire variables:
  - tree_survival
  - avg_severity
  - dist_to_edge
  - litter cover
- veg cover & shrub biomass
- forb & shrub biomass
- tree diversity & biomass

<br>

### Collapse variables using PCA

#### Categories for collapsing:

**Fire**

- *avg_severity* - mean severity of pixels within the site, based on MTBS fire severity data (0-3 = unburned-high severity)
- *dist_to_edge* - distance from the site center to the burn boundary (negative for sites outside the burn boundary)
- *tree_survival* - proportion of live trees (%)
- *litter_cover*

```{r fire PCA}

# Fire

dat_fire <- habitat %>% 
  select(site, severity, avg_pixel, dist_to_edge, tree_survival)

fire_pca <- prcomp(dat_fire[3:5],
                 scale = TRUE) # scales the data

#summary(fire_pca) # PC1 describes 85% of the variance

colors <- c('darkgreen', 'orange', 'red')

ggbiplot(fire_pca, groups = dat_fire$severity) +
  scale_color_manual(name = 'Severity', values = colors) +
  xlim(-3, 3) +
  theme_classic()

fire_scores <- data.frame(fire_pca$x) %>% 
  bind_cols(site_meta) %>% 
  select(site, severity, PC1) %>% 
  rename(fire_score = PC1)

```

**Biomass**  

- *tree_bm* - density of adult tree biomass (live or dead) (kg/m^2^)
- *shrub_bm* - density of live shrub biomass (kg/m^2^)
- *grass_bm* - density of live grass biomass (kg/m^2^)
- *forb_bm* - density of live forb biomass (kg/m^2^)

```{r biomass PCA}

# Biomass

dat_biomass <- habitat %>% 
  select(site, severity, tree_bm, shrub_bm, grass_bm, forb_bm)

biomass_pca <- prcomp(dat_biomass[3:6],
                 scale = TRUE) # scales the data

#summary(biomass_pca) # PC1 describes 54% of the variance

ggbiplot(biomass_pca, groups = dat_fire$severity) +
  scale_color_manual(name = 'Severity', values = colors) +
  xlim(-3, 3) +
  theme_classic()

biomass_scores <- data.frame(biomass_pca$x) %>% 
  bind_cols(site_meta) %>% 
  select(site, severity, PC1) %>% 
  rename(biomass_score = PC1)

```

**Vegetation diversity**

- *tree_H* - Shannon diversity of adult trees
- *shrub_H* - Shannon diversity of adult trees
- *forb_H* - Shannon diversity of adult trees

```{r diversity PCA}

# Diversity

dat_diversity <- habitat %>% 
  select(site, severity, tree_H, shrub_H, forb_H)

diversity_pca <- prcomp(dat_diversity[3:5],
                 scale = TRUE) # scales the data

#summary(diversity_pca) # PC1 describes 58% of the variance

ggbiplot(diversity_pca, groups = dat_fire$severity) +
  scale_color_manual(name = 'Severity', values = colors) +
  xlim(-3, 3) +
  theme_classic()

diversity_scores <- data.frame(diversity_pca$x) %>% 
  bind_cols(site_meta) %>% 
  select(site, severity, PC1) %>% 
  rename(diversity_score = PC1)

```

**Miscellaneous** (don't collapse)

- *tree_density* - density of trees (live or dead) (indv/m^2^)
- *veg_cover* - total proportion of vegetation cover (all species/life-forms - can sum to >100) (both transects) (%)
- *litter_cover* - proportion of litter cover (%)
- *cwd_volume* - density of coarse woody debris volume (m^3^/m^2^)

```{r, eval=FALSE}

# Make data frame with new PC1 scores

pc_scores <- full_join(fire_scores, biomass_scores) %>% 
  full_join(diversity_scores)

# Export data

habitat_metrics <- full_join(habitat, pc_scores)

write_csv(habitat_metrics,'output_data/habitat_metrics.csv', col_names = TRUE)

```

```{r eval=FALSE}

rmarkdown::render('code/07_habitat_pca.Rmd', output_file = '../docs/07_habitat_pca.html')

```
