---
title: "NMDS"
author: "Kate Culhane"
output:
  html_document:
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = '..')
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.width=7, fig.height=5)
```

<br>

```{r packages and data}

# Load packages
library(vegan)
library(goeveg)
library(pairwiseAdonis)
library(ggvegan)
library(ggrepel)
library(tidyverse)

# Load data

matrix_df <- read_csv('output_data/species_matrix.csv') %>% 
  mutate(severity = factor(severity, levels = c('unb', 'mod', 'high')))

matrix_adj_df <- read_csv('output_data/species_matrix_adj.csv') %>% 
  mutate(severity = factor(severity, levels = c('unb', 'mod', 'high')))

site_meta <- read_csv('raw_data/field_data/site_metadata.csv') %>% 
  mutate(severity = factor(severity, levels = c('unb', 'mod', 'high')))

veg_metrics <- read_csv('output_data/veg_metrics.csv') %>% 
  mutate(severity = factor(severity, levels = c('unb', 'mod', 'high')))

smamm_meta <- read_csv('raw_data/lit_data/mammal_traits.csv')

```

```{r data wrangling}

# Transform data frames into matrices

matrix <- as.matrix(matrix_df[-c(1,2)])
rownames(matrix) <- matrix_df$site

matrix_adj <- as.matrix(matrix_adj_df[-c(1,2)])
rownames(matrix_adj) <- matrix_adj_df$site

matrix_hell <- decostand(matrix, 'hellinger')

matrix_sqrt <- sqrt(matrix)

# Severity data frame
sevs <- select(matrix_df, site, severity)

```

## Community structure

[*Overview of data transformation and standardization methods*](https://chrischizinski.github.io/SNR_R_Group/2016-08-10-Data-Transformations)

```{r include=FALSE}

# NMDS with relative abundance data (2 dimensions)
set.seed(17)
ord <- metaMDS(matrix_hell,
               distance = 'bray',
               autotransform = FALSE,
               k = 3, # number of dimensions
               try=50)
ord$stress # 0.096

```

```{r, fig.height=4, results='hide',fig.keep='all', fig.show="hold", out.width="50%"}

# Stress plot
stressplot(ord)

# Scree plot to check stress per number of dimensions
dimcheckMDS(matrix_hell, distance = "bray", autotransform = FALSE, k = 10, trymax = 20)

```

**NMDS plot showing community structure across the three fire severities:** 

- Species codes:
    - SPBE = California ground squirrel
    - GLSA = northern flying squirrel
    - TAAM = yellow-pine chipmunk
    - TAQU = long-eared chipmunk
    - TASE = shadow chipmunk
    - NEFU = dusky-footed woodrat
    - PEMA = North American deer mouse
    - PEBO = brush mouse
    - PETR = Pinyon mouse
    - REME = western harvest mouse
    - SOTR = Trowbridge's shrew
    
```{r}

cols <- c('darkgreen', 'orange', 'red')
par(mar = c(4,4,1,1))
plot(ord, display = 'species', type = 'n')
points(ord, display = 'sites',
       cex = 1.5, pch = 19, col = cols[sevs$severity])
ordiellipse(ord, groups = sevs$severity,
            label = TRUE, col = cols, lwd = 2)
text(ord, display = 'species',
     cex = 0.6)

```

- Raw data standardized using the Hellinger method
- Uses Bray-Curtis dissimilarity matrix
- 3-dimensional ordination (2 dim shown)
- Stress = 0.09
- Appears to be fewer sites because some of them exactly overlap

<br>

### ANOSIM

```{r include=FALSE}

# Attach severity data to species matrix
matrix_hell
attach(sevs)

# Compare community similarity across fire severities using ANOSIM
smamm_ano <- anosim(matrix_hell, severity)

```

```{r fig.height=4, fig.width=4}

plot(smamm_ano)
summary(smamm_ano)

```

```{r include=FALSE}

# Compare community dispersion across fire severities using betadisper
smamm_dist <- vegdist(matrix_hell)
smamm_bd <- betadisper(smamm_dist, severity)

smamm_bd
plot(smamm_bd)
boxplot(smamm_bd)

permutest(smamm_bd)
anova(smamm_bd)
# p > 0.05 for both --> no difference in dispersion across fire severities
# Satisfies the assumption for adonis

```

<br>

### Adonis

```{r}

# Compare community similarity across fire severities using adonis
smamm_ado <- adonis(matrix_hell ~ severity,
                    permuatations = 999,
                    method = 'bray')
smamm_ado

# Pairwise Adonis
smamm_ado_pw <- pairwise.adonis(matrix_hell, severity,
                                sim.method = 'bray',
                                p.adjust.m = 'bonferroni')
smamm_ado_pw

```

- Adonis is more robust because it is less sensitive to dispersion
- Used betadisper to determine that dispersion was similar across treatments in order to satisfy assumptions of Adonis

<br>

## Habitat drivers

```{r include=FALSE}

# Data frame with only the variables from the model
model_vars <- veg_metrics %>% 
  select(site, severity,
         litter_cover, live_tree_density, soft_cwd_volume,
         shrub_cover, forb_grass_cover, pca_score)

# Fit environmental variables
fit <- envfit(ord, model_vars[,3:8], permutations = 1000)

# NMDS plot with arrows for environmental variables
par(mar = c(4,4,1,1))
plot(ord, diplay = 'sites', type = 'n')
points(ord, display = 'sites',
       cex = 1, pch = 19, col = cols[sevs$severity])
ordiellipse(ord, groups = sevs$severity, # ellipse with std error around centroids
            label = FALSE, col = cols, lwd = 2)
plot(fit,
     col = 'black', lwd = 3, cex = 0.8)
text(ord, display = 'species',
     cex = 0.7, col = 'grey30')

```

```{r, include=FALSE}

# Data frames with point and vector informations

site_points <- fortify(ord) %>% 
  filter(Score == 'sites') %>% 
  full_join(sevs, by = c('Label' = 'site')) %>% 
  dplyr::rename(group = severity)


spp_points <- fortify(ord) %>% 
  filter(Score == 'species')

vectors <- fortify(fit) %>% 
  mutate(xend=NMDS1*1.4,
         yend=NMDS2*1.4,
         xlab = case_when(
           Label == 'shrub_cover' ~ xend-0.1,
           Label == 'litter_cover' ~ xend-0.3,
           Label == 'live_tree_density' ~ xend-0.32,
           Label == 'soft_cwd_volume' ~ xend-0.1,
           Label == 'forb_grass_cover' ~ xend-0.3,
           Label == 'pca_score' ~ xend+0.24
         ),
         ylab = case_when(
           Label == 'shrub_cover' ~ yend+0.08,
           Label == 'litter_cover' ~ yend-0.04,
           Label == 'live_tree_density' ~ yend,
           Label == 'soft_cwd_volume' ~ yend-0.08,
           Label == 'forb_grass_cover' ~ yend+0.08,
           Label == 'pca_score' ~ yend+0.03
         ),
         lab = case_when(
           Label == 'shrub_cover' ~ 'Shrub cover',
           Label == 'litter_cover' ~ 'Litter cover',
           Label == 'live_tree_density' ~ 'Tree density',
           Label == 'soft_cwd_volume' ~ 'Soft CWD',
           Label == 'forb_grass_cover' ~ 'Forb+grass cover',
           Label == 'pca_score' ~ 'PC1[T+S+L]'
         ))

# Data frame with ellipse information

plot.new()
ord_el <- ordiellipse(ord, groups = sevs$severity)

veganCovEllipse<-function (cov, center = c(0, 0), scale = 1, npoints = 100) 
  {
    theta <- (0:npoints) * 2 * pi/npoints
    Circle <- cbind(cos(theta), sin(theta))
    t(center + scale * t(Circle %*% chol(cov)))
  }

df_ell <- data.frame()

for(g in levels(site_points$group)){
  df_ell <- rbind(df_ell, cbind(as.data.frame(with(site_points[site_points$group==g,],
                  veganCovEllipse(ord_el[[g]]$cov,ord_el[[g]]$center,ord_el[[g]]$scale)))
                                ,group=g))
}

```

```{r}
  
# Plot
ggplot() +
  geom_path(data=df_ell, # severity ellipses
            aes(x=NMDS1, y=NMDS2, color=group),
            size=0.6) +
  geom_point(data=site_points, # site points
             aes(x=NMDS1, y=NMDS2, color=group, shape=group),
             size=1.8) +
  geom_segment(data=vectors[1:5,], # insig veg arrows
               aes(x=0, y=0, xend=xend, yend=yend), 
               arrow=arrow(length=unit(0.02, 'npc'), type='closed'),
               size=0.6) +
  geom_segment(data=vectors[6,], # sig veg arrow
               aes(x=0, y=0, xend=xend, yend=yend), 
               arrow=arrow(length=unit(0.02, 'npc'), type='closed'),
               size=0.6, linetype=5) +
  geom_text(data=vectors[1:5,], # insig veg arrow labels
            aes(x=xlab, y=ylab, label=lab), size=4) +
  geom_text(data=vectors[6,], # sig veg arrow label
            aes(x=xlab, y=ylab, label=lab), size=4, parse=TRUE) +
  geom_label(data=site_points, # NMDS information box
             aes(x=0.63, y=-0.85,
                 label='NMDS in 3 dim\n2 dim displayed\nstress = 0.096'),
             size=3.5) +
  scale_color_manual(values=c('#4AB793', '#DCA827', '#D85F2B')) +
  scale_shape_manual(values=c(19,17,15)) +
  scale_x_continuous(limits=c(-1.2,0.9)) +
  scale_y_continuous(limits=c(-1,0.6)) +
  theme_classic() +
  theme(panel.border=element_rect(fill=NA),
        axis.text.x=element_blank(), axis.text.y=element_blank(),
        legend.position='none') +
  coord_fixed(clip='off')

ggsave('figs/fig4b.png', plot=last_plot(),
       width=3.5, height=3, units='in')

```

<br>

## Traits

```{r include=FALSE}

# Data frame with mammal traits
traits <- read_csv('raw_data/lit_data/mammal_traits_NEW.csv') %>% 
  arrange(species)

# Fit traits
trait_fit <- envfit(ord, traits[,c(4:6)], permutations = 1000, display = 'sp')

# NMDS plot with arrows for traits
par(mar = c(4,4,1,1))
plot(ord, diplay = 'sites', type = 'n')
points(ord, display = 'sites',
       cex = 1, pch = 19, col = cols[sevs$severity])
ordiellipse(ord, groups = sevs$severity, # ellipse with std error around centroids
            label = FALSE, col = cols, lwd = 2)
plot(trait_fit,
     col = 'black', lwd = 3, cex = 0.8)
text(ord, display = 'species',
     cex = 0.7, col = 'grey30')

vectors2 <- fortify(trait_fit) %>% 
  mutate(xend=NMDS1,
         yend=NMDS2,
         lab = case_when(
           Label == 'nestsBurrow' ~ 'Burrow',
           Label == 'nestsHollow' ~ 'Hollow',
           Label == 'nestsTree' ~ 'Tree',
           Label == 'guildInsectivore' ~ 'Insectivore',
           Label == 'guildOmnivore'~ 'Omnivore',
           Label == 'guildHerbivore' ~ 'Herbivore',
           Label == 'foragArboreal' ~ 'Arboreal',
           Label == 'foragGround' ~ 'Ground',
           Label == 'foragScansorial' ~ 'Scansorial'
         ))

vectors3 <- vectors %>% 
  mutate(
    xlab = case_when(
      Label=='soft_cwd_volume'~xlab+0.22,
      Label=='pca_score'~xlab,
      TRUE~xlab
    ),
    ylab = case_when(
      Label=='soft_cwd_volume'~ylab,
      Label=='pca_score'~ylab,
      TRUE~ylab
    )
  )

```

```{r}

# Plot

ggplot() +
  geom_path(data=df_ell, aes(x=NMDS1, y=NMDS2, color=group),
            size=0.6) +
  geom_segment(data=vectors2, aes(x=0, y=0, xend=NMDS1, yend=NMDS2),
               arrow=arrow(length=unit(0.02, 'npc'), type='closed'), size=0.6) +
  geom_text(data=vectors2, aes(x=NMDS1, y=NMDS2, label=lab),
            size=4) +
  geom_segment(data=vectors3[c(3,6),], aes(x=0, y=0, xend=xend, yend=yend,),
               arrow=arrow(length=unit(0.02, 'npc'), type='closed'),
               size=0.6, linetype=5) +
  geom_text(data=vectors3[c(3,6),], aes(x=xlab, y=ylab, label=lab),
            size=4, fontface='bold') +
  geom_label(data=site_points, aes(x=0.82, y=-0.7,
                                   label='NMDS in 3 dim\n2 dim displayed\nstress = 0.09'),
             size=4) +
  scale_color_manual(values=c('#4AB793', '#DCA827', '#D85F2B')) +
  scale_shape_manual(values=c(19,17,15)) +
  scale_x_continuous(limits=c(-1.6,1.15)) +
  scale_y_continuous(limits=c(-0.9,0.71)) +
  theme_classic() +
  theme(panel.border=element_rect(fill=NA),
        axis.text.x=element_blank(), axis.text.y=element_blank(),
        legend.position='none') +
  coord_fixed(clip='off')

```

```{r, include=FALSE}

# Species labels

spp_points2 <- spp_points %>% 
  full_join(smamm_meta[c(1,5)], by = c('Label'='species')) %>% 
  mutate(xdat=NMDS1*0.61,
         ydat=NMDS2*0.7)

ggplot() +
  geom_text_repel(data=spp_points2, aes(x=-1, y=ydat, label=common_name),
                  direction='y', segment.alpha=0, hjust=0,
                  size=3.5, color='grey40', point.padding = 0) +
  scale_x_continuous(limits=c(-1.6,1.15)) +
  scale_y_continuous(limits=c(-0.9,0.7)) +
  theme_classic() +
  theme(panel.border=element_rect(fill=NA),
        axis.text.x=element_blank(), axis.text.y=element_blank(),
        legend.position='none') +
  coord_fixed(clip='off')

ggplot() +
  geom_text_repel(data=spp_points2, aes(x=xdat, y=-0.7, label=common_name),
                  direction='x', segment.alpha=0, vjust=0,
                  size=3.5, color='grey40', angle=90, point.padding = 0) +
  scale_x_continuous(limits=c(-1.6,1.15)) +
  scale_y_continuous(limits=c(-0.9,0.7)) +
  theme_classic() +
  theme(panel.border=element_rect(fill=NA),
        axis.text.x=element_blank(), axis.text.y=element_blank(),
        legend.position='none') +
  coord_fixed(clip='off')

```

```{r eval=FALSE}
rmarkdown::render('code/09_nmds.Rmd', output_file = '../docs/09_nmds.html')
```
