---
title: "Functional Diversity"
author: "Kate Culhane"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = '..')
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.width=5, fig.height=3)
```

<br>

### Functional trait data

Trait data compiled from:

- Audubon and Peterson field guides and Journal of Mammalogy species accounts (diet and foraging mode traits)
- PanTHERIA: a species-level database of life history, ecology, and geography of extant and recently extinct mammals - 2005 taxonomy (colonization ability traits)
  - Missing species data filled with genus-level data
- Field measurements (body size traits)
- PHYLACINE 1.2: The Phylogenetic Atlas of Mammal Macroecology (misc. analyses)

```{r packages & data}

# Load packages

library(tidyverse)
library(ggmosaic)
library(FD)
library(dunn.test)
library(ggpubr)
library(gridExtra)
library(npmv)

select <- dplyr::select
group_by <- dplyr::group_by
summarize <- dplyr::summarize
rename <- dplyr::rename
mutate <- dplyr::mutate

# Load data

smamms <- read_csv('output_data/unique_smamms.csv') %>% 
  mutate(severity = factor(severity, levels = c('unb', 'mod', 'high')))

captures <- read_csv('output_data/smamms.csv') %>% 
  mutate(severity = factor(severity, levels = c('unb', 'mod', 'high')))

effort <- read_csv('output_data/effort.csv') %>% 
  mutate(severity = factor(severity, levels = c('unb', 'mod', 'high')))

site_meta <- read_csv('raw_data/field_data/site_metadata.csv') %>% 
  mutate(severity = factor(severity, levels = c('unb', 'mod', 'high')))

traits <- read_csv('raw_data/lit_data/mammal_traits.csv') 

phylacine <- read_csv('raw_data/lit_data/PHYLACINE.csv')

pantheria <- read_csv('raw_data/lit_data/PanTHERIA.csv') 

```

```{r data wrangling}

# Data frame containing only species information

species <- smamms %>% 
  group_by(species) %>% 
  filter(row_number()==1) %>% 
  select(species, family, binomial, binomial_short)

# Data frame with abundance per species, per site

abundance <- smamms %>% 
  group_by(site, severity, species, binomial) %>% 
  summarize(n = length(species)) %>% 
  full_join(effort) %>% 
  mutate(cpue = n/effort*300)

# Update binomials in PHYLACINE dataset

phylacine$binomial <- gsub('_', ' ', phylacine$binomial)

# Filter PHYLACINE data

phylacine_mass <- semi_join(phylacine, species, by = 'binomial') %>% 
  select(family, genus, species, binomial, mass)

phylacine_diet <- semi_join(phylacine, species, by = 'binomial') %>% 
  select(family, genus, species, binomial, diet_plant, diet_invert, diet_vert) %>% 
  gather(key = 'diet_type', value = 'diet_percent', 5:7) %>% 
  filter(diet_percent != 0) %>% 
  mutate(diet_type = factor(diet_type, c('diet_plant', 'diet_invert', 'diet_vert')))

```

```{r mass data wrangling}

# Data frame of the average mass for each unique individual

capture_mass <- captures %>% 
  select(site, severity, indID, binomial, weight, age) %>% 
  filter(!is.na(indID), # filter for captures only
         (age == 'A' | binomial == 'Neotamias amoenus')) %>% # filter for only adult animals (all NEAM were subadults, unfortunately)
  group_by(indID, site, severity, binomial) %>% 
  summarize(avg_mass = mean(weight, na.rm = TRUE)) %>% 
  ungroup()

```

### Individual Components of Functional Diversity

#### Body mass

```{r}

# Data frame of the average mass for each species

species_mass <- capture_mass %>% 
  group_by(binomial) %>% 
  summarize(spp_avg = mean(avg_mass, na.rm = TRUE))

write_csv(species_mass, 'output_data/species_mass.csv', col_names = TRUE)

```

#### Biomass

```{r biomass data wrangling}

# Combined data frame with the average mass for each unique individual, with NAs filled in with species averages

capture_biomass <- full_join(capture_mass, species_mass, by = 'binomial') %>% 
  mutate(mass = case_when(
    !is.na(avg_mass) ~ avg_mass,
    is.na(avg_mass) ~ spp_avg
  )) %>% 
  select(site, severity, binomial, mass)

# Data frame of total biomass at each site

biomass_byspp <- capture_biomass %>% 
  group_by(site, severity, binomial) %>% 
  summarize(biomass = sum(mass)) %>% 
  ungroup() %>% 
  mutate(severity = factor(severity, c('unb', 'mod', 'high')))

biomass <- biomass_byspp %>% 
  group_by(site, severity) %>% 
  summarize(biomass_total = sum(biomass)) %>% # calculate total biomass at each site, summed across species
  mutate(severity = factor(severity, c('unb', 'mod', 'high')))

```

**Increase of total biomass was not significant across the fire severity gradient:** 

```{r biomass boxplot}

# Box and whisker plot of the mean of total site biomass across severities

colors <- c('darkgreen', 'orange', 'red')
xlabs <- c('Unburned', 'Moderate Severity', 'High Severity')

ggplot(biomass, aes(x = severity, y = biomass_total)) +
  geom_boxplot(aes(fill = severity)) +
  scale_fill_manual(values = colors) +
  theme_minimal() +
  labs(x = 'Fire Severity',
       y = 'Total biomass (g)') +
  theme(legend.position = 'NA') +
  scale_x_discrete(labels = xlabs)

```

```{r total biomass across severities kruskal-wallis, eval=FALSE}


biomass %>% 
  group_by(severity) %>% 
  summarize(
    biomass_mean = mean(biomass_total),
    biomass_sd = sd(biomass_total)
  )

# Does the total biomass at each site vary across a fire severity gradient?

# Explore the data distribution and variance

ggplot(biomass, aes(x = biomass_total)) +
  geom_histogram() +
  facet_wrap(~severity)

ggplot(biomass, aes(sample = biomass_total)) +
  geom_qq() +
  facet_wrap(~severity)

variances <- biomass %>% 
  group_by(severity) %>% 
  summarize(variance = var(biomass_total))

# Histograms look a little funky because sample size is so small
# qq-plots look weird and swoopy, pluass the mod distribution is really disjunct
# n=9 for each severity, so cannot use CLT to justify normality

# Variances are WAY different (>4x)
# Cannot satisfy assumptions of ANOVA



# Test for significant differences in means

# Kruskal Wallis
kruskal.test(biomass_total ~ severity, data = biomass)
# p>0.05, so there is no sig difference in means

```

- Mean total site biomass increased with fire severity, but the differences between treatments was not significant (Kruskal-Wallis test, (chi square (2) = 2, p = 0.37, n = 9 for each).
- Results are the same using PHYLACINE and PanTHERIA body size data
- Outliers are sites that had a lot of ground squirrels

<br>

**Total biomass by species across the fire severity gradient:** 

```{r biomass column graph, fig.width=6, fig.height=4}

# Column graph of biomass by species for each fire severity

ggplot(biomass_byspp, aes(x = severity, y = biomass, fill = binomial)) +
  geom_col() +
  theme_minimal() +
  labs(x = 'Fire Severity',
       y = 'Biomass (g)',
       fill = 'Species')

```

- Total biomass summed across sites for each severity, with the proportion contributed by each species.
- Increases in biomass were mainly driven by *P. maniculatus* and *O. beecheyi*.
- Strange that even though the mean biomass did not change between severities, the total biomass was much higher for mod and high severities (skewed by ground squirrel outliers)

<br>

#### Diet

**Abundance by feeding guild across the fire severity gradient:** 
 
```{r feeding guild col graph}

# Column graph of feeding guild by fire severity

guild <- full_join(abundance, traits) %>% 
  full_join(traits) %>% 
  group_by(severity, guild) %>% 
  summarize(n = sum(cpue))

ggplot(guild, aes(x = severity, y = n, fill = guild)) +
  geom_col() +
  theme_minimal() +
  labs(x = 'Fire severity',
       y = 'Number of individuals',
       fill = 'Feeding guild')

```

- All three treatments were dominated by generalist species.
- Categorical feeding guild (herbivore, omnivore, carnivore, insectivore) was not used in functional diversity calculations

<br> 

**Community-wide daily food consumption for each diet item:**

```{r diet data frame}

# Data frame of TOTAL abundance per species (summed across sites): use to order binomial factors

total_spp_n <- smamms %>% 
  group_by(binomial) %>% 
  summarize(n = length(binomial)) %>% 
  arrange(-n)

# Data frame of abundance of each diet type by site and species, weighted by biomass from field data

diet <- full_join(biomass_byspp, traits, by = 'binomial') %>% 
  select(severity, site, binomial, biomass, seeds, fruits, veg, fungi, inverts, verts) %>% 
  mutate(dfc = (10.5*(biomass^0.507))/14, # daily food consumption, calulated as FMR/ME
         seeds = seeds*dfc/100, # weight diet proportions by dfc for each indv.
         fruits = fruits*dfc/100,
         veg = veg*dfc/100,
         fungi = fungi*dfc/100,
         inverts = inverts*dfc/100,
         verts = verts*dfc/100) %>% 
  group_by(severity, site, binomial) %>% 
  summarize(seeds = sum(seeds),
            fruits = sum(fruits),
            veg = sum(veg),
            fungi = sum(fungi),
            inverts = sum(inverts),
            verts = sum(verts)) %>%
  gather(key = diet_type, value = n, 4:9) %>% 
  ungroup() %>% 
  mutate(binomial = factor(binomial, total_spp_n$binomial),
         diet_type = factor(diet_type, c('veg', 'seeds', 'fruits', 'fungi', 'inverts', 'verts')),
         severity = factor(severity, c ('unb','mod','high')))

# Data frame summed by severity, species, and diet type

diet_sev <- diet %>% 
  group_by(severity, binomial, diet_type) %>% 
  summarize(total = sum(n))

# Data frame summed by site and diet type

diet_site <- diet %>%
  group_by(site, diet_type, severity) %>% 
  summarize(total = sum(n))

```

```{r diet npmv, eval=FALSE}

# Data frame summed by site and diet type, wide format

diet_spread <- diet_site %>% 
  spread(key = diet_type, value = total)
  
# Non-parametric comparison of multivariate samples (package npmv)
# Non-parametric equivalent of MANOVA

nonpartest(fruits|fungi|inverts|seeds|veg|verts ~ severity,
           diet_spread,
           plots = FALSE)

# p>0.05 so there is no difference between groups

```

```{r diet jitter plots, fig.height=4, fig.width=6}

# Jitter plots

ggline(diet_site, x = 'severity', y = 'total',
                      add = c('mean_se', 'jitter'),
                      facet.by = 'diet_type', scale = 'free') +
  labs(x = 'Severity',
       y = 'DFC')

```

- DFC for each diet item did not vary across the fire severity gradient.
- Differences in the community-wide DFC (in grams/day, summed across individuals of species at each site) between treatments for each diet item were not significant (nonparametric one-way ANOVA-type multivariate test F(4.2, 50.5) = 2.47, p = 0.054).
- DFC scales allometrically with body mass and is equal to FMR/ME
- Biomass was calculated from field measurements

<br>

**Mosaic plot of community diet across fire severities:** 

```{r diet mosaic plot, eval=FALSE, fig.height=3.5, fig.width=7, include=FALSE}

# Mosaic plot

ggplot(data = diet_sev) +
  geom_mosaic(aes(weight = total, x = product(diet_type), fill = binomial),
              color = 'black',
              lwd = 0.5,
              offset = 0) + 
  scale_y_productlist(breaks=c(0, 0.25, 0.50, 0.75, 1),
                      labels=c('0', '25', '50', '75', '100')) +
  labs(x = 'DFC proportion by diet item (%)',
       y = 'DFC proportion by species (%)',
       fill = 'Species') +
  theme_minimal() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle = 90, vjust = 0, hjust = 1)) +
  facet_grid(cols = vars(severity),
             scale = 'free',
             labeller = as_labeller(c('unb' = 'Unburned',
                                      'mod' = 'Low-Moderate',
                                      'high' = 'High')))

```

- Proportion of total daily food consumption (summed across sites) of each type of diet item by species for each severity. Although the overall proportion of DFC for each type of diet item remained consistent between treatments, the species consuming each diet item varied.
- DFC was calculated similarly to above

<br>

#### Foraging mode

**Abundance by foraging height across fire severity treatments:** 

```{r foraging height col graph}

# Column graph of foraging height by fire severity

height <- abundance %>% 
  full_join(traits, by = 'binomial') %>% 
  group_by(severity, arboreal) %>% 
  summarize(n = sum(cpue))

ggplot(height, aes(x = severity, y = n, fill = arboreal)) +
  geom_col() +
  theme_minimal() +
  labs(x = 'Fire severity',
       y = 'Number of individuals',
       fill = 'Foraging Height')

```

- All severities were dominated by scansorial species.
- Values:
  - 0 = fully terrestrial
  - 1 = scansorial - lower canopy
  - 2 = scansorial - upper canopy
  - 3 = fully arboreal (no representatives in our data)
  
<br>

**Abundance by degree of fossoriality across fire severity treatments:** 

```{r fossoriality col graph}

# Column graph of degree of fossoriality by fire severity

foss <- abundance %>% 
  full_join(traits, by = 'binomial') %>% 
  group_by(severity, fossorial) %>% 
  summarize(n = sum(cpue))

ggplot(foss, aes(x = severity, y = n,
                             fill = fossorial)) +
  geom_col() +
  theme_minimal() +
  labs(x = 'Fire severity',
       y = 'Number of individuals',
       fill = 'Degree of Fossoriality')

```

- All severities were dominated by non-fossorial species.
- Values:
  - 0 = non-fossorial
  - 1 = semi-fossorial (sometimes forages underground)
  - 2 = semi-fossorial (sometimes forages and caches food underground)
  - 3 = fully fossorial (no representatives in our data)

<br>

#### Activity time

**Abundance by activity time across fire severity treatments.** 

```{r activity time col graph}

# Column graph of activity time by fire severity

activity <- abundance %>% 
  full_join(traits, by = 'binomial') %>% 
  group_by(severity, activity) %>% 
  summarize(n = sum(cpue))

ggplot(activity, aes(x = severity, y = n,
                                     fill = activity)) +
  geom_col() +
  theme_minimal() +
  labs(x = 'Fire severity',
       y = 'Number of individuals',
       fill = 'Activity Time')

```

- All severities were dominated by nocturnal species.
- This may be misleading because traps were primarily open at night.

<br>

### Functional Diversity Indices 

**Index calculation:**

1. Uses Gower dissimilarity matrix (can use ordered, categorical, and numerical variables)
    - integrates variable weights
2. Conducts PCA on the distance matrix and uses the resulting PC axes as 'traits' to compute FD indices
3. Calculates indices based on equations in (Laliberte & Legendre, 2010)

**Metrics:**

- Functional richness (FRic)
- Functional divergence (FDiv)
- Functional evenness (FEve)
- Functional dispersion (FDis)

**Functional traits included:**

- Body size
- Diet
  - Seeds
  - Fruits
  - Vegetation (other)
  - Fungi
  - Invertebrates
  - Vertebrates
- Foraging mode
  - Foraging height
  - Degree of fossoriality
- Colonization ability
  - Home range
  - Litter size
  - Annual litters 
  - Longevity
  - Age at sexual maturity

<br>

```{r FD indices 3, include=FALSE}

# Data frame of functional traits

trait <- inner_join(species_mass, traits, by = 'binomial') %>% 
  inner_join(pantheria, by = 'binomial') %>% 
  select(binomial, spp_avg, seeds, fruits, veg, fungi, inverts, verts, arboreal, fossorial, home_range_indv, litter_size, annual_litters, max_longevity, age_sexualmaturity) %>% 
  arrange(binomial)

spp <- trait$binomial # vector of just species

trait <- trait %>% 
  select(-binomial) # remove species column
rownames(trait) <- spp

# Matrix of abundances by species

abun_temp <- abundance %>% 
  ungroup() %>% 
  select(site, binomial, n) %>% 
  spread(binomial, n) %>%
  ungroup() %>% 
  replace(., is.na(.), 0) # replace NAs with 0

abun <- as.matrix(abun_temp[-1])
rownames(abun) <- abun_temp$site

# Calculate functional diversity using FD package

weight <- c(0.25, # weight for body mass
            0.0417, 0.0417, 0.0417, 0.0417, 0.0417, 0.0417, # weight for diet categories
            0.125, 0.125, # weight for foraging mode
            0.05, 0.05, 0.05, 0.05, 0.05) # weight for weediness attributes

smamm_fd <- dbFD(trait, abun, weight)

smamm_fd$FRic # functional richness
smamm_fd$FDiv # functional divergence
smamm_fd$FEve # functional evenness
smamm_fd$FDis # functional dispersion

# Data frame with values for the four indices at each site

indices <- data_frame(site = abun_temp$site,
                       FRic = smamm_fd$FRic,
                       FDiv = smamm_fd$FDiv,
                       FEve = smamm_fd$FEve,
                       FDis = smamm_fd$FDis) %>% 
  full_join(site_meta[3:4])

```

**Functional diversity did not vary across the fire severity gradient:**

```{r FD jitter plots 3, fig.width=6, fig.height=4}

# Jitter plots of the mean of FRic across severities
fric_jitter <- ggline(indices, x = 'severity', y = 'FRic',
                      add = c('mean_se', 'jitter'),
                      xlab = 'Fire Severity',
                      ylab = 'FRic')

# Jitter plots of the mean of FDiv across severities
fdiv_jitter <- ggline(indices, x = 'severity', y = 'FDiv',
                      add = c('mean_se', 'jitter'),
                      xlab = 'Fire Severity',
                      ylab = 'FDiv')

# Jitter plots of the mean of FEve across severities
feve_jitter <- ggline(indices, x = 'severity', y = 'FEve',
                      add = c('mean_se', 'jitter'),
                      xlab = 'Fire Severity',
                      ylab = 'FEve')

# Jitter plots of the mean of FDis across severities
fdis_jitter <- ggline(indices, x = 'severity', y = 'FDis',
                      add = c('mean_se', 'jitter'),
                      xlab = 'Fire Severity',
                      ylab = 'FDis')

# All four plots
grid.arrange(fric_jitter, fdiv_jitter, feve_jitter, fdis_jitter)

```

```{r FD by severity 3, eval=FALSE}

indices %>% 
  dplyr::group_by(severity) %>% 
  dplyr::summarise(
    FEve_mean = mean(FEve, na.rm = TRUE)
  )

# Does functional diversity vary across a fire severity gradient?

# Explore the data distribution

ggplot(indices, aes(x = FRic)) +
  geom_histogram() +
  facet_wrap(~severity)
ggplot(indices, aes(x = FDiv)) +
  geom_histogram() +
  facet_wrap(~severity)
ggplot(indices, aes(x = FEve)) +
  geom_histogram() +
  facet_wrap(~severity)
ggplot(indices, aes(x = FDis)) +
  geom_histogram() +
  facet_wrap(~severity)
# Histograms are weird
# n=9 for each severity, so cannot use CLT to justify normality

indices %>% 
  group_by(severity) %>% 
  summarize(
    var_fric = var(FRic, na.rm = TRUE),
    var_fdiv = var(FDiv, na.rm = TRUE),
    var_feve = var(FEve, na.rm = TRUE),
    var_fdis = var(FDis)
  )
# Variance for FEve and FRic can satisfy assumption for ANOVA (<4x), but might as well run KW for everything



# Test for significant differences in means

# FRic: Kruskal Wallis
kruskal.test(FRic ~ severity, data = indices)
dunn.test(indices$FRic, indices$severity, method = 'bonferroni')

# FDiv: Kruskal Wallis
kruskal.test(FDiv ~ severity, data = indices)

# FEve: Kruskal Wallis
kruskal.test(FEve ~ severity, data = indices)
dunn.test(indices$FEve, indices$severity, method = 'bonferroni')

# FDis: Kruskal-Wallis
kruskal.test(FDis ~ severity, data = indices)

```

- Calculated values for each site are displayed for four indices of functional diversity: functional richness (FRic), functional divergence (FDiv), functional evenness (FEve), and functional dispersion (FDis). Means and standard deviations are shown for the three fire severity treatments: unburned (unb), low-moderate severity (mod), and high severity (high). Functional diversity indices were calculated using four ecologically relevant traits: body size, diet, foraging mode, and colonization ability.
- Differences in FRic are significant by Kruskal-Wallis (p = 0.01925)
  - Only different between high sev and unburned (Dunn's p = 0.0092)
- I don't account for phylogenetic correlation in any way

```{r eval=FALSE}

# Export csv with biomass and FD per site

FD_csv <- full_join(biomass, indices)

write_csv(FD_csv, 'output_data/FD_biomass.csv', col_names = TRUE)

```

```{r eval=FALSE}

rmarkdown::render('code/10_functional_diversity.Rmd', output_file = '../docs/10_functional_diversity.html')

```
