---
title: "Multivariate GLM"
author: "Kate Culhane"
output:
  html_document:
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = '..')
knitr::opts_chunk$set(warning = FALSE, message = FALSE,
                      fig.height=3.5, fig.width=5)
```

<br>

```{r packages and data, include=FALSE}

# Load packages
library(mvabund)
library(vegan)
library(qpcR) # for calculating AIC weights
library(tidyverse)

# Load data

matrix_df <- read_csv('output_data/species_matrix.csv') %>% 
  mutate(severity = factor(severity, levels = c('unb', 'mod', 'high')))

veg_metrics <- read_csv('output_data/veg_metrics.csv') %>% 
  mutate(severity = factor(severity, levels = c('unb', 'mod', 'high')))

site_centroids <- read_csv('output_data/site_centroids.csv')

effort <- read_csv('output_data/effort.csv') %>% 
  mutate(offset = effort/300)

```

```{r include=FALSE}

# Principle Components of Neighborhood Matrix (PCNM)

site_pcnm <- pcnm(dist(site_centroids[,2:3]))

pcnm_vectors <- as.data.frame(site_pcnm$vectors) %>% 
  cbind(site_centroids$sites_aea.site)%>% 
  rename(site = 'site_centroids$sites_aea.site',
         pcnm_1 = PCNM1) %>% 
  select(site, pcnm_1)

# All predictor variables

vars <- veg_metrics %>% 
  full_join(pcnm_vectors) %>% 
  select(-tree_survival)

```

Resources:

- [Introduction from Warton](http://eco-stats.blogspot.com/2012/03/introducing-mvabund-package-and-why.html)
- [Tutorial](https://www.dropbox.com/s/unwd412a9wbkrl0/mvabundTute.R?dl=0)
- [Powerpoint about mvabund](http://www.seec.uct.ac.za/sites/default/files/image_tool/images/330/SEEC%20Toolbox%20mvabund.pdf) (video lecture is also useful)
- [Differences between `anova` and `summary`](https://stats.stackexchange.com/questions/59879/logistic-regression-anova-chi-square-test-vs-significance-of-coefficients-ano)
- [Wald vs. score test statistic](https://stats.idre.ucla.edu/other/mult-pkg/faq/general/faqhow-are-the-likelihood-ratio-wald-and-lagrange-multiplier-score-tests-different-andor-similar/)

```{r include=FALSE}

# Create mvabund object using species matrix
dat <- full_join(matrix_df, vars)
spp <- mvabund(dat[,3:13])

```

```{r include=FALSE, eval=FALSE}

# Compare relationship between means and variances
meanvar.plot(spp) # looks like a linear increase...
# need to use appropriate GLM that assumes a similar mean-variance relationship --> negative binomial

plot(data=dat, spp~severity)

```

<br>

## GLM~severity~

- Tests for differences in multivariate community structure among the three fire severity categories
- i.e. *Does treatment have an effect on assemblage?*
- Analogous to the ANOSIM and Adonis
- `anova` determines which predictors are the strongest contributors by sequentially adding them to the restricted (less variables) model (for test=score)
  - Used the score statistic since means are small because of rare species (this would decrease power when using the Wald statistic)

```{r echo=TRUE}

glm <- manyglm(data=dat, spp ~ severity + pcnm_1,
               family='negative_binomial', # to account for mean-variance relationship
               offset = effort$offset) # offset accounts for trapping effort

plot(glm) # check for model fit

an <- anova(glm,
            test='score', # allows for small means due to rare species
            cor.type='shrink', # accounts for correlation between variables
            nBoot=999) # number of bootstrap iterations
an$table

```

<br>

### Indicator species

- Tests univariate (species) differences by severity
- Same model, different p-value calculation

```{r echo=TRUE}

an2 <- anova(glm, test='score', cor.type='shrink', nBoot=999, p.uni='adjusted')
an2

```

<br>

## GLM~veg~

- Tests for differences in multivariate community structure according to the vegetation variables
- i.e. *Which environmental variables are most strongly associated with assemblages?*
- `summary` determines which predictors are the strongest contributors when all other predictors have been fitted to the restricted (less variables) model (true for for test=score)

**Hypothesis predictor variables:**

- PC1~TSL~
  - Live tree density
  - Shrub cover
  - Litter cover
- Soft coarse woody debris (CWD)
- Forb/grass cover

**Model with ALL hypothesis predictor variables**

```{r echo=TRUE}

glm2 <- manyglm(data=dat, spp ~ pca_score + soft_cwd_volume + forb_grass_cover + pcnm_1,
               family='negative_binomial', # to account for mean-variance relationship
               offset=effort$offset)# offset accounts for trapping effort

plot(glm2) # check for model fit

sum <- summary(glm2,
               test='score', # allows for small means due to rare species
               cor.type='shrink', # accounts for correlation between variables
               nBoot=999) # number of bootstrap iterations
sum

```

<br>

### AIC selection

```{r echo=FALSE}

# Create data frame with all possible variable combinations
veg <- c('soft_cwd_volume', 'forb_grass_cover', 'pca_score', 'pcnm_1')
cbind.fill <- function(...){
    nm <- list(...) 
    nm <- lapply(nm, as.matrix)
    n <- max(sapply(nm, nrow)) 
    do.call(cbind, lapply(nm, function (x) 
        rbind(x, matrix(,n-nrow(x),ncol(x))))) 
}
var_combos <- data.frame(cbind.fill(combn(veg,1), combn(veg,2), combn(veg,3), combn(veg,4)))

# Calculate AIC for each combination
model_aic <- data_frame(vars=character(),
                        K=numeric(),
                        aic=numeric())
for (i in 1:ncol(var_combos)) {
  vars <- levels(var_combos[,i])
  vars_dat <- as.matrix(select(dat,vars))
  K <- length(vars)
  aic <- manyglm(spp~vars_dat, family='negative_binomial', offset=effort$offset)$AICsum
  new_combo <- data_frame(vars=c(paste(vars,collapse=" + ")),
                          K=K,
                          aic=aic)
  model_aic <- rbind(model_aic, new_combo)
}

aic_arrange <- model_aic %>% 
  arrange(aic)

weights <- akaike.weights(aic_arrange$aic)

model_selection <- cbind(aic_arrange,
                         delta_aic=weights$deltaAIC,
                         weight=weights$weights) %>% 
  mutate(aic=round(aic,2),
         delta_aic=round(delta_aic,2),
         weight=round(weight,2))

head(model_selection, 6)
```

**Final model**

Vegetation variables included:

- PC1~TSL
- Soft CWD

```{r echo=TRUE}

glm3 <- manyglm(data=dat, spp ~ pca_score + soft_cwd_volume,
               family='negative_binomial', # accounts for mean-variance relationship
               offset=effort$offset) # accounts for trapping effort

plot(glm3) # check for model fit

sum <- summary(glm3,
               test='score', # allows for small means due to rare species
               cor.type='shrink', # accounts for correlation between variables
               nBoot=999) # number of bootstrap iterations
sum

```

<br>

## GLM~trait~

Resources:

- [Introduction by Warton](https://rpubs.com/dwarton/68823)
- [Paper using similar methods](https://royalsocietypublishing.org/doi/10.1098/rsos.200076#d3e468)

```{r include=FALSE}

# Data wrangling
trait_new <- read.csv('raw_data/lit_data/mammal_traits_NEW.csv') %>% 
  arrange(species)
dat2 <- as.data.frame(dat)
L_dat <- dat2[c(3:13)] %>% 
  mutate_each(funs(if(is.numeric(.)) as.integer(.) else .))
rownames(L_dat) <- dat$site
R_dat <- dat2[c(2,19:20)]
rownames(R_dat) <- dat$site
traits <- trait_new[c(3:6)]
rownames(traits) <- trait_new$species

```

<br>

### manyGLM method

- No option for including an offset to account for trapping effort, so used relative abundance data instead
  - Using relative instead of absolute abundance improved model fit by decreasing heteroscedasticity of residuals
- `anova` tests whether the environment-trait interactions predict community structure 

```{r echo=TRUE}

fc_glm <- traitglm(L_dat, R_dat[-1], traits[-1],
                   composition=TRUE, # relative abundance
                   method='manyglm',
                   family='negative.binomial') # accounts for mean-variance relationship

plot(fc_glm) # check for model fit

fc_anova <- anova(fc_glm)
fc_anova

```

<br>

### LASSO penalty method

- Uses the LASSO penalty to do model selection
- Removes interaction terms that do not decrease BIC
- Cannot use `anova` with this method
- Shows which environment-trait interactions are most important for determining community structure

```{r echo=TRUE}

fc <- traitglm(L_dat, R_dat[-1], traits[-1],
               composition=TRUE, # relative abundance
               method='glm1path', # LASSO penalty
               family='negative.binomial') # accounts for mean-variance relationship

plot(fc) # check model fit

```

```{r echo=FALSE, fig.height=4, fig.width=3}

fourth <- as_tibble(cbind(trait=rownames(fc$fourth),fc$fourth)) %>% 
  pivot_longer(cols=2:3,
               names_to='veg') %>% 
  separate(trait, sep=5, into=c('group','trait')) %>% 
  mutate(
    value=as.numeric(value),
    vegetation=case_when(
      veg=='pca_score'~'PC1[T+S+L]',
      veg=='soft_cwd_volume'~'Soft CWD'
    ),
    vegetation=factor(vegetation, levels=c('PC1[T+S+L]', 'Soft CWD')),
    group=case_when(
      group=='nests'~'Nesting habit',
      group=='guild'~'Feeding guild',
      group=='forag'~'Foraging mode'
    ),
    group=factor(group,levels=c('Feeding guild','Foraging mode','Nesting habit')),
    trait=factor(trait, levels=c('Insectivore','Omnivore','Herbivore',
                                 'Ground','Scansorial','Arboreal',
                                 'Burrow','Hollow','Tree'))
    )

fourth %>% 
  ggplot(aes(x=vegetation, y=trait, fill=value)) +
  geom_tile() +
  scale_fill_gradient2(low='darkred', high='darkblue', limits=c(-0.4,0.4)) +
  scale_x_discrete(labels=parse(text=c('PC1[T+S+L]','Soft~CWD'))) +
  facet_grid(rows=vars(group), scales='free_y', space='free', switch='y') +
  theme_minimal() +
  theme(panel.border = element_rect(fill=NA),
        panel.spacing = unit(0,'cm'),
        axis.text.x = element_text(angle=35, vjust=1.3, hjust=1.2,
                                   face='bold', color='black'),
        axis.ticks = element_line(),
        strip.placement = 'outside',
        strip.background.y = element_rect(fill=NA),
        strip.text = element_text(face='bold'),
        legend.key.height = unit(0.5,'in'),
        axis.title = element_text(size=10),
        legend.key.width = unit(0.1,'in'),
        legend.title = element_blank(),
        legend.box.spacing = unit(0.05,'in'),
        panel.grid = element_blank()) +
  labs(x='Vegetation variable', y='Small mammal trait')

ggsave('figs/fig6.png', plot = last_plot(),
       width = 2.5, height = 3.7, units = 'in')

```

- **Figure 6**
- Heat map showing estimated fourth corner coefficients

```{r eval=FALSE, include=FALSE}
rmarkdown::render('code/10_mvglm.Rmd', output_file = '../docs/10_mvglm.html')
```

