---
title: "Phylogenetic Diversity"
author: "Kate Culhane"
date: "January 31, 2019"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = '..')
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.width=5, fig.height=3)
```

<br>

```{r packages & data}

# Load necessary packages

library(tidyverse)
library(ape)
library(picante)
library(lefse)
library(geiger)
library(ggpubr)

# Load necessary data

smamms <- read_csv('output_data/unique_smamms.csv') %>% 
  mutate(severity = factor(severity, levels = c('unb', 'mod', 'high'))) %>% 
  mutate(binomial_new = gsub(' ', '_', binomial)) # update binomial format for analyses

# Create species matrix with binomials as the column names

matrix <- smamms %>% 
  group_by(site, severity, binomial_new) %>% 
  tally() %>% # count individuals per site
  spread(binomial_new, n) %>% # reformat into matrix
  ungroup() %>% 
  replace(., is.na(.), 0) %>% # replace NAs with 0
  arrange(severity)

```

### Phylogenetic data

- Tree from PHYLACINE 1.2: The Phylogenetic Atlas of Mammal Macroecology
- 11 species total
  - 10 species from order Rodentia
  - 1 species from order Soricomorpha

```{r phylogeny}

# Phylogeny

tree <- read.nexus('raw_data/lit_data/small_phylogeny.nex')

phylo <- tree[[1]]

# Trim tree

species <- unique(smamms$binomial_new)

pruned <- drop.tip(phylo, phylo$tip.label[-match(species, phylo$tip.label)])

par(mar = c(1,1,1,1))
plot(pruned)

```

<br>

### Weighted Faith's PD

```{r PA PD, include=FALSE}

# FAITH'S PD - P/A

# Convert species matrix to presence/absence

matrix_pa <- matrix[, -2]
for(i in 2:ncol(matrix_pa)) {
  matrix_pa[which(matrix_pa[,i] > 0), i] <- 1
}

# Calculate Faith's PD for each site

pd <- data.frame(matrix_pa$site, pd(matrix_pa[,-1], pruned, include.root = F))
names(pd)[1] <- 'site'

# Calculate PD of a community containing all taxa in regional phylogeny

allTaxaCom <- matrix_pa[1, -1]
allTaxaCom[1,] <- 1
pd(allTaxaCom, pruned, include.root = F) # highest PD is 364.26

```


```{r weighted faiths pd}

# Calculate weighted Faith's PD for each site

matrix2 <- matrix %>% 
  as.data.frame(.) %>% 
  mutate(Peromyscus_maniculatus = as.numeric(Peromyscus_maniculatus)) %>% 
  filter(site != 'HICORN' & site != 'HISQUE' & site != 'HITEDD' & site != 'HIWALK') # filter out sites with only one species (PEMA)

sites <- matrix2$site

matrix3 <- matrix2 %>% 
  select(-site, -severity)

rownames(matrix3) <- sites

weighted_pd <- as.data.frame(weighted.faith(pruned, matrix3)) %>% 
  mutate(site = rownames(.)) %>% 
  rename(PD = 'weighted.faith(pruned, matrix3)') %>% 
  select(site, PD)



# Calculate PD for sites with 1 species (P. maniculatus)
    # Equal to the branch length for P. maniculatus

PEMA <- 'Peromyscus_maniculatus'

pema_prune <- drop.tip(pruned, pruned$tip.label[-match(PEMA, pruned$tip.label)])

pema_pd <- pema_prune$edge.length

# Update PD data frame

pd_indices <- full_join(weighted_pd, matrix[1:2]) %>% 
  mutate(PD = replace_na(PD, pema_pd))

```

**Weighted Faith's PD decreased across the fire severity gradient:** 
    
```{r FD jitter plots, echo=FALSE, warning=FALSE}

# Jitter plot of the mean of PD across severities

ggline(pd_indices, x = 'severity', y = 'PD',
       add = c('mean_se', 'jitter'),
       xlab = 'Fire Severity',
       ylab = 'Weighted Faiths PD')

```

```{r PD by severity, eval=FALSE}

# Does phylogenetic diversity vary across a fire severity gradient?

# Explore the data distribution

ggplot(pd_indices, aes(x = PD)) +
  geom_histogram() +
  facet_wrap(~severity)

pd_indices %>% 
  group_by(severity) %>% 
  summarize(far_pd = var(PD, na.rm = TRUE))

# Histogram is weird
# n=9 for each severity, so cannot use CLT to justify normality
# Variances are NOT close enough for ANOVA



# Test for significant differences in means

# Kruskal Wallis and post-hoc pairwise testing (Dunn's test)

kruskal.test(PD ~ severity, data = pd_indices)

dunn.test(pd_indices$PD, pd_indices$severity,
          method = 'bonferroni') # Bonferroni correction

# ANOVA

aov(PD ~ severity, data = pd_indices)

```

- Significant by both Kruskal-Wallis and ANOVA (p<0.05).
- For communities with only one species, Faith's PD (weighted & unweighted) equals the branch length for that one species
    - When the 4 communities with one species are left out, there's still a significant differnece between severities, but high overlaps greatly with both unb and mod
- Outliers
    - High severity: HISHRU and HIBADR are high because they had Peromyscus, Tamias, and Otospermophilus species
    - Low-mod severity: MIXMIX is low because it only had two Peromyscus species
    
<br>

### Removing Sorex from PD estimates

```{r weighted faiths pd --sorex, eval=FALSE}

# Calculate weighted Faith's PD for each site

matrix2 <- matrix %>% 
  as.data.frame(.) %>% 
  mutate(Peromyscus_maniculatus = as.numeric(Peromyscus_maniculatus)) %>% 
  filter(site != 'HICORN' & site != 'HISQUE' & site != 'HITEDD' & site != 'HIWALK') %>% # filter out sites with only one species (PEMA)
  select(-Sorex_trowbridgii) %>% # remove SOTR
  filter(site != 'GRNCAR' & site != 'GRNQUA' & site != 'GRNSNO') # filter out sites with only one species when SOTR is excluded

sites <- matrix2$site

matrix3 <- matrix2 %>% 
  select(-site, -severity)

rownames(matrix3) <- sites

weighted_pd_nosorex <- as.data.frame(weighted.faith(pruned, matrix3)) %>% 
  mutate(site = rownames(.)) %>% 
  rename(PD = 'weighted.faith(pruned, matrix3)') %>% 
  select(site, PD)

# Fill with PD for sites with one species

pd_indices_nosorex <- full_join(weighted_pd_nosorex, matrix[1:2]) %>% 
  mutate(PD = replace_na(PD, pema_pd))

```

**Weighted Faith's PD with *Sorex trowbridgii* removed did not decrease across the fire severity gradient:**

```{r FD jitter plots --sorex, eval=FALSE}

# Jitter plot of the mean of PD across severities

ggline(pd_indices_nosorex, x = 'severity', y = 'PD',
       add = c('mean_se', 'jitter'),
       xlab = 'Fire Severity',
       ylab = 'Weighted Faiths PD')

```

```{r PD by severity --sorex, eval=FALSE}

# Does phylogenetic diversity vary across a fire severity gradient?


# Explore the data distribution

ggplot(pd_indices2, aes(x = PD)) +
  geom_histogram() +
  facet_wrap(~severity)

pd_indices2 %>% 
  group_by(severity) %>% 
  summarize(far_pd = var(PD, na.rm = TRUE))

# Histogram is weird
# n=9 for each severity, so cannot use CLT to justify normality

# Variances are close enough for ANOVA



# Test for significant differences in means

# Kruskal Wallis and post-hoc pairwise testing (Dunn's test)

kruskal.test(PD ~ severity, data = pd_indices2)
# p = 0.3217

# ANOVA

pd_aov <- aov(PD ~ severity, data = pd_indices2)
summary(pd_aov) # p = 0.45

```

- Not significant by both Kruskal-Wallis and ANOVA (p > 0.3).

```{r eval=FALSE}

# Write csv with PD values for each site

pd_csv <- full_join(pd_indices, pd_indices_nosorex, by = c('site', 'severity')) %>% 
  rename(PD = PD.x,
         PD_nosorex = PD.y)

write_csv(pd_csv, 'output_data/PD.csv', col_names = TRUE)

```

```{r eval=FALSE}

rmarkdown::render('code/11_phylogenetic_diversity.Rmd', output_file = '../docs/11_phylogenetic_diversity_NORARE.html')

```
