---
title: "Failed GLMMs"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}

library(lme4) # for GLMMs
library(multcomp) # for GLMM Tukey
library(piecewiseSEM) # for GLMM r-squared
library(jtools) # for GLM plotting
library(tidyverse)

# Load data

smamms <- read_csv('output_data/unique_smamms.csv') %>% 
  mutate(severity = factor(severity, levels = c('unb', 'mod', 'high')))

captures <- read_csv('output_data/smamms.csv') %>% 
  mutate(severity = factor(severity, levels = c('unb', 'mod', 'high')))

effort <- read_csv('output_data/effort.csv') %>% 
  mutate(severity = factor(severity, levels = c('unb', 'mod', 'high')))

mass <- read_csv('output_data/species_mass.csv')

site_meta <- read_csv('raw_data/field_data/site_metadata.csv') %>% 
  mutate(severity = factor(severity, levels = c('unb', 'mod', 'high')))

```


### GLMM comparing total abundance among fire severities

response ~ severity + random(block) + offset(effort)

Abundance response variables (total, PEMA, and SOTR abundance): Poisson
Other response variables (biomass): Gamma

Richness, evenness: Gaussian???

R2: https://jonlefcheck.net/2013/03/13/r2-for-linear-mixed-effects-models/
Offsets: http://environmentalcomputing.net/interpreting-coefficients-in-glms/
Communicating results: http://environmentalcomputing.net/mixed-models-3/
Using GLMs instead of LMs: https://besjournals.onlinelibrary.wiley.com/doi/full/10.1111/2041-210X.12552#support-information-section
Choosing a family: https://datavoreconsulting.com/post/count-data-glms-choosing-poisson-negative-binomial-zero-inflated-poisson/

```{r}

# Plot mean vs. variance
abundance %>% 
  group_by(severity) %>% 
  summarize(mean(n),
            var(n))
# Overdispersed data --> use negative binomial distribution

# GLMMs
glmm <- glmer.nb(data=abundance, n ~ severity + (1|block) + offset(offset))

# Null model for comparison
glmm_null <- glmer.nb(data=abundance, n ~ 1 + (1|block) + offset(offset))

# Check assumptions
plot(glmm)
ggqqplot(residuals(glmm))

# Model stats
summary(glmm)
coef(glmm)
ci <- confint(glmm, level=0.95, method='Wald')

# Compare to null model
anova(glmm, glmm_null, test='Chisq')

# Test each severity category
summary(glht(glmm, mcp(severity="Tukey")))

# Model fit
rsquared(glmm)




# Create GLM
glm <- glm.nb(data=abundance, n ~ severity + offset(offset))
glm_null <- glm.nb(data=abundance, n ~ 1 + offset(offset))

glm <- glm(data=abundance, n ~ severity + offset(offset), family=quasipoisson)

# Predict model mean and SE
dat <- tibble(severity=c('unb','mod','high'))
pred <- predict(glm, newdata=dat, type='link', se=TRUE)
exp <- exp(pred$fit)
se_up <- exp(pred$fit + 1.96 * pred$se.fit)
se_low <- exp(pred$fit - 1.96 * pred$se.fit)
modfit <- tibble(dat,exp,se_up,se_low)




# Test for fit of Poisson
1-pchisq(summary(glm)$deviance, summary(glm)$df.residual)
plot(glm)
summary(glm)
anova(glm, glm_null, test='Chisq')
summary(glht(glm, mcp(severity="Tukey")))

# Check for under/overdispersion
E2 <- resid(glm, type = "pearson")
N  <- nrow(abundance)
p  <- length(coef(glm))   
sum(E2^2) / (N - p)

# Plot
ggplot() +
  geom_jitter(data=abundance, aes(x=severity, y=n, color=severity, shape=severity),
              width=0.2, size=2) +
  scale_color_manual(values=cols) +
  scale_shape_manual(values=c(1,2,0)) +
  geom_point(data=modfit, aes(x=severity, y=exp),
             size=2) +
  geom_errorbar(data=modfit, aes(x=severity, ymin=se_low, ymax=se_up),
                width=0.2, size=0.6) +
  theme_classic() +
  labs(x = 'Fire Severity',
       y = 'Total small mammal abundance') +
  theme(legend.position = 'NA') +
  scale_x_discrete(labels = xlabs)
```

<br>

### GLMM comparing PEMA abundance among fire severities

```{r}

# GLMM
glmm <- glmer(data=pema, n_pema ~ severity + (1|block) + offset(offset),
              family=poisson)

# Null model for comparison
glmm_null <- glmer(data=pema, n_pema ~ 1 + (1|block) + offset(offset),
                  family=poisson)

# Check assumptions
plot(glmm)
ggqqplot(residuals(glmm))

# Model stats
summary(glmm)
coef(glmm)
ci <- confint(glmm, level=0.95, method='Wald')

# Compare to null model
anova(glmm, glmm_null, test='Chisq')

# Test each severity category
summary(glht(glmm, mcp(severity="Tukey")))

# Model fit
rsquared(glmm)

# Create GLM without random effect (to test for fit of Poisson, and for plotting)
glm <- glm(data=pema, n_pema ~ severity + offset(offset),
           family=poisson)
glm_null <- glm(data=pema, n_pema ~ 1 + offset(offset))

# Test for fit of Poisson
1-pchisq(summary(glm)$deviance, summary(glm)$df.residual)
plot(glm)
summary(glm)
anova(glm, glm_null, test='Chisq')
summary(glht(glm, mcp(severity="Tukey")))

# Plot raw data and model predictions
pred <- make_predictions(glm, pred='severity', data=pema)

effect_plot(glm, pred=severity, data=pema,
            plot.points=TRUE, jitter=.2)

ggplot() +
  geom_jitter(data=pema, aes(x=severity, y=n_pema, color=severity, shape=severity),
              width=0.2, size=2) +
  scale_color_manual(values=cols) +
  scale_shape_manual(values=c(1,2,0)) +
  geom_point(data=pred, aes(x=severity, y=n_pema),
             size=3) +
  geom_errorbar(data=pred, aes(x=severity, ymin=ymin, ymax=ymax),
                width=0.2, size=1) +
  theme_classic() +
  labs(x = 'Fire Severity',
       y = 'Deer mouse abundance') +
  theme(legend.position = 'NA') +
  scale_x_discrete(labels = xlabs)

```

<br>

### GLMM comparing SOTR abundance among fire severities

```{r}

# GLMM
glmm <- glmer(data=sotr, n_sotr ~ severity + (1|block) + offset(offset),
              family=poisson)

# Null model for comparison
glmm_null <- glmer(data=sotr, n_sotr ~ 1 + (1|block) + offset(offset),
                  family=poisson)

# Check assumptions
plot(glmm)
ggqqplot(residuals(glmm))

# Model stats
summary(glmm)
coef(glmm)
ci <- confint(glmm, level=0.95, method='Wald')

# Compare to null model
anova(glmm, glmm_null, test='Chisq')

# Test each severity category
summary(glht(glmm, mcp(severity="Tukey")))

# Model fit
rsquared(glmm)

# Create GLM without random effect (to test for fit of Poisson, and for plotting)
glm <- glm(data=sotr, n_sotr ~ severity + offset(offset),
           family=poisson)
glm_null <- glm(data=sotr, n_sotr ~ 1 + offset(offset))

glm <- glm(data=sotr, n_sotr ~ severity + offset(offset),
           family=quasipoisson)

# Test for fit of Poisson
1-pchisq(summary(glm)$deviance, summary(glm)$df.residual)
plot(glm)
summary(glm)
anova(glm, glm_null, test='Chisq')
summary(glht(glm, mcp(severity="Tukey")))

# Plot raw data and model predictions
pred <- make_predictions(glm, pred='severity', data=sotr)

effect_plot(glm, pred=severity, data=sotr,
            plot.points=TRUE, jitter=.2)


ggplot() +
  geom_jitter(data=pema, aes(x=severity, y=n_pema, color=severity, shape=severity),
              width=0.2, size=2) +
  scale_color_manual(values=cols) +
  scale_shape_manual(values=c(1,2,0)) +
  geom_point(data=pred, aes(x=severity, y=n_pema),
             size=3) +
  geom_errorbar(data=pred, aes(x=severity, ymin=ymin, ymax=ymax),
                width=0.2, size=1) +
  theme_classic() +
  labs(x = 'Fire Severity',
       y = 'Deer mouse abundance') +
  theme(legend.position = 'NA') +
  scale_x_discrete(labels = xlabs)
```